---
title: Get your stats result in a table easily
author: Sebastian Sauer
draft: TRUE
date: '2017-11-21'
slug: get-your-stats-result-in-a-table-easily
categories:
  - rstats
tags:
  - rstats
  - writing
---

```{r echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)

```

Having computed some staticis, one would like to display them. Either in a figure, on in a table, that's the two typical ways.

Let's explore some helper functions to get your stats to a table easily.

A nice overview on packages can be found [here](https://stackoverflow.com/questions/5465314/tools-for-making-latex-tables-in-r).

Let's have a quick glance at these steps. We will use the dataset `flights` of the package `nycflights13`.

```{r}
data(flights, package = "nycflights13")
```


This post will be tidyverse-driven:

```{r}
library(tidyverse)
library(skimr)
```

Let's shorten the dataset:

```{r}
flights %>% 
  select(air_time, distance, arr_delay) %>% 
  sample_n(1000) -> flights_short
```

Let's compute a summary:



```{r}
mosaic::favstats(~arr_delay, data = flights_short)
```


# Descriptive statistics

## `ApaTables` ad `apaStyle`

A number of suitables packages exist, for example, checkout `apaTables`, `apaStyle`, or `xtable`.

 See:
```{r eval = FALSE}
apaStyle::apa.descriptives(data = flights[c("arr_delay", "distance")],
                           variables = c("arr_delay", "distance"))
```

With the result: "Word document succesfully generated in: /Users/sebastiansauer/Documents/RRRing/data_se"


## `kable`

```{r}
library(knitr)
```

```{r}
kable
```




## `xtable`
Remember to use the chunk option `results = "asis"`.

```{r}
library(xtable)
```


```{r results = "asis"}
options(xtable.comment = FALSE)  # prevent verbose comments
flights_short %>% 
  head %>% 
  xtable %>% 
  print(type = "html")
```




## `stargazer`
Remember to use the chunk option `results = "asis"`.

```{r}
library(stargazer)
```


```{r results = "asis"}
flights_short %>% 
  skim %>% 
  stargazer(type = "html")
```


# Linear models

```{r}
lm1 <- lm(arr_delay ~ origin + distance, data = flights)
```

```{r results = "asis"}
stargazer(lm1, type = "html")
```

