---
title: Compute effect sizes with R. A primer.
author: Sebastian Sauer
date: '2017-11-21'
slug: compute-effect-sizes-with-r-a-primer
categories:
  - rstats
tags:
  - rstats
---



```{r echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)

```


A typical "cook book recipe" for doing data analysis is an applied stats course is:

1. report descriptive statistics
2. plot some nice diagrams
3. test hypothesis
4. report effect sizes


Let's have a quick glance at these steps. We will use the dataset `flights` of the package `nycflights13`.

```{r}
data(flights, package = "nycflights13")
```


This post will be tidyverse-driven:

```{r}
library(tidyverse)
library(skimr)
```


# Report descriptive statistics


```{r}
flights %>% 
  skim
```

Done!

Oh, wait, "subgroup" analysis is some often asked for task.

```{r}
flights %>% 
  group_by(origin) %>% 
  skim
```


# Bonus: Use mosaic

The package mosaic is another beginer friendly approach to data analysis.

```{r}
library(mosaic)
```


```{r}
mosaic::inspect(flights)
```


# Bonus: Get the results printed in nice table

