---
title: Plot many ggplot diagrams using nest() and map()
author: ''
date: '2018-12-05'
slug: plot-many-ggplot-diagrams-using-nest-and-map
categories:
  - rstats
tags:
  - rstats
  - tidyverse
  - plotting
editor_options: 
  chunk_output_type: console
---

```{r knitr-steup, echo = FALSE}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "100%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =   .618,
  fig.show = "hold",
  size = "tiny"
)
```



At times, it is helpflu to plot a multiple of related diagrams, such as a scatterplot for each subgroup. As always, there a number of ways of doing so in R. Specifically, we will make use of `ggplot2`.

```{r}
library(tidyverse)
library(glue)
```


```{r}
data(mtcars)
```



```{r}
d <- mtcars %>% 
  rownames_to_column(var = "car_names")
```


Is `d` a tibble`

```{r}
is_tibble(d)
```

What is it?


```{r}
class(d)
```


Okay, let's make a tibble out of it:

```{r}
d <- as_tibble(d)
class(d)
```


# Way 1: using facets

One simple way is to plot several facets according to the grouping variable:


```{r}
d %>% 
  ggplot() +
  aes(x = hp, y = mpg) +
  geom_point() +
  facet_wrap(~ cyl)
```


# Way 2: using `nest()` and `map2()`


First, we nest the data frame:

```{r}
d_nested <- 
  d %>% 
  group_by(cyl) %>% 
  nest()

d_nested
```

Note that in the column `data` there is the data of mtcars - broken down for each group. That's why we have three lines.


Second, we map the data to ggplot:


```{r}
d_plots <- 
  d_nested %>% 
  mutate(plot = map2(data, cyl, ~ ggplot(data = .x, aes(x = hp, y = mpg)) +
                      ggtitle(glue("Number of Cylinder: {.y}")) +
                      geom_point()))

d_plots
```

Finally, we print it:


```{r}
print(d_plots$plot)
```



Note that we need `map2()` because in the data (`data`), there is no information on the number of cylinders. Hence, we need to hand over a second vector with the cylinder information. If two vectors serving as input for `map()`, we need `map2()`.


# Way 3: Deprecated `do`


There are of course other ways to achieve what we just explored. For example, good ol' for-loops are out there. However, here we can make use of R's beautiful vectorization capabilities. In addition, `dplyr::do()` is a similar way to map list elements to a function. However, this function ~~is~~will probably get deprecated:


```{r echo=FALSE}
blogdown::shortcode('tweet', '719542847045636096')
```


In addition, `nest()` bears th advantage, that the (processed plotting) data is nicely stored in a data frame.


For the sake of completeness:

```{r}
d_plots2 <- d %>% 
  group_by(cyl) %>% 
  dplyr::do(plot = {ggplot(data = ., aes(x = hp, y= mpg)) + geom_point() + ggtitle(paste0("Cylinders: ", .$cyl))})

d_plots2
```


```{r}
print(d_plots2$plot)
```

