---
title: Change standard theme of ggplot
author: ''
date: '2018-10-10'
draft: FALSE
slug: change-standard-theme-of-ggplot
categories:
  - dataviz
  - rstats
tags:
  - plotting
  - ggplot2
editor_options: 
  chunk_output_type: console
---

```{r echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  1,
  fig.show = "hold",
  size = "tiny"
)
```



`ggplot2` is customizeable. Frankly, one can change a heap of details - not everything probably, but a lot. Of course, one can add a theme to the ggplot call, in order to change the theme. However, a more catch-it-all approach would be to change the standard theme of ggplot itself. In this post, we'll investigate this option.


Load some data and the right packages:

```{r}
data(mtcars)
library(tidyverse)
```


Here's the standard theme of ggplot, let's have a look at it
```{r}
theme_gray
```


Now let's make a copy of it, to be on the save side:

```{r}
theme_gray2 <- theme_gray
```

Here's a demo plot, assuming we want to change some aspects of the theme, such as color or font type.


```{r}
ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_jitter()
```




The standard theme is `theme_gray`, which can easily be updated in this way:



```{r}
theme_update(text = element_text(family = "Times",
                            colour = "red"))
```





Here comes the new (updated) theme:


```{r}
ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_jitter()
```


Alternatively, we could just apply our new theme:

```{r}
theme_test <- theme_grey() +
  theme(text = element_text(family = "Times",
                            color = "blue"))


ggplot(mtcars, aes(x = cyl, y = mpg)) +
  geom_jitter() + 
  theme_test
```

