---
title: Power calculation for the general linear model
author: ''
date: '2018-07-24'
slug: power-calculation-for-the-general-linear-model
categories:
  - rstats
tags:
  - rstats
---


Before conducting an experiment, one should compute the power - or, preferably, estimate the precision of the expected results. There are numerous way to achieve this, here's one using the R package `pwr`.


# Package pwr

```{r}
library(pwr)
```


The workhorse function here is `pwr.f2.test`. Note that `f2` refers to the effect size $f^2$ (see [here](https://en.wikipedia.org/wiki/Effect_size#Cohen's_%C6%922)), defined as:

$$f^2 = \frac{R^2}{1-R^2}$$.


See for details of the function its help page:
```{r}
help("pwr.f2.test")
```


```
pwr.f2.test(u = NULL, v = NULL, f2 = NULL, sig.level = 0.05, power = NULL)
```

Note that `u` is the number of predictors ($k$) minus 1, that's the numerator degrees of freedom (df). The intercepts counts a predictor, too in this case!

`v` is the denominator df, defined as $n-p$, where $n$ is the sample size (see [here](https://math.stackexchange.com/questions/617735/multiple-regression-degrees-of-freedom-f-test) for some details on df).

By the way, if you wonder, what df actually are, read [this](https://www.quora.com/How-can-I-explain-degrees-of-freedom-simply-and-clearly) nice little explanation.

Let's look at one example. In a power analysis before the experiment, one states the desired level of power, and the expected effect size, for example:

```{r}
# example: k=3 predictors, n=100 observations, R^2 = .1, sig.level = .05 (default)

pwr.f2.test(u = 2,
            v = 97,
            f2 = .1/.9,
            power = NULL)
```

Given this values, we will have a power of about 84% (if all assumptions are met).


The other way round: How large should the sample be to get a power of .8, given $R^2=.^1$?

```{r}
# example: k=3 predictors, power = .8, R^2 = .1

pwr.f2.test(u = 2,
            f2 = .1/.9,
            power = .8,
            v = NULL)

```

`v = 87`, ie, $n=v+p=87+3=90$.



