---
title: Plot columns repeatedly
author: ''
date: '2018-11-02'
slug: plot-columns-repeatedly
categories:
  - rstats
tags:
  - plotting
  - rstats
editor_options: 
  chunk_output_type: console
---


Suppose you have a large number of columns of a dataframe, and you want to plot each column -- say a histogram for each column.

This post shows some ways of achieving this.

Let's take the mtcars dataset as an example.

```{r}
data(mtcars)
```


We will use the tidyverse approach:

```{r}
library(tidyverse)
```


# Way 1


```{r}
mtcars %>% 
  map2(., names(.), ~ {ggplot(data = data_frame(.x), aes(x = .x)) + 
      geom_histogram() +
      labs(x= .y)})
```


Some explanations:

- First, we take the dataset `mtcars`.
- Then, we map a function (ie., `ggplot()`) to each column of `mtcars`, but we also parse the names of `mtcars`.
- `ggplot()` likes dataframes, but `map()` serves lists/vectors, so we have to enshrine each vector to a dataframe using `data_frame()`.
- The data comes from the first list (`mtcars`), that's where `.x` comes from (or refers to).
- The names come from the second lsit (`names(mtcars)`), that's where `.y` points to.


# Way 2

A maybe more simple is this:


```{r}
mtcars %>% 
  gather(key = item, value = value) %>% 
  ggplot() +
  aes(x = value) +
  geom_density() +
  facet_wrap(~ item, ncol = 2, scales = "free")
```



# Test if column is normally distributed before doing anything else

Suppose we want to check whether a column is nicely normally distributed before plotting. That's one way to checking that:


```{r}
mtcars %>% 
  map(~ shapiro.test(.x)) %>% 
  map("p.value") %>% 
  keep(. > .05)
```

