---
title: Binning and recoding with R - some recommendations
author: ''
date: '2018-08-09'
slug: binning-and-recoding-with-r-some-recommendations
categories:
  - rstats
tags:
  - rstats
---


*Recoding* means changing the levels of a variable, for instance changing "1" to "woman" and "2" to "man". *Binning* means aggregating several variable levels to one, for instance aggregating the values From "1.00 meter" to "1.60 meter" to "small_size".

Both operations are frequently necessary in practical data analysis. In this post, we review some methods to accomplish these two tasks.


```{r echo  = FALSE}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  0.4,  #0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)

```



Let's load some example data:

```{r}
data(tips, package = "reshape2")
```


Some packages:
```{r}
library(mosaic)
```


One nice way is using the function `case_when()` from the `tidyverse` community. Consider this example:

```{r}
tips$tip_gruppe <- case_when(
  tips$tip < 2 ~ "scrooge",
  tips$tip < 4 ~ "ok",
  tips$tip < 8 ~ "generous",
  TRUE ~ "in love"
)

```



Wait, `case_when` is pipe-friendly, see:

```{r}
tips <- tips %>% 
  mutate(tip_gruppe = case_when(
    tip < 2 ~ "scrooge",
    tip < 4 ~ "ok",
    tip < 8 ~ "generous",
    TRUE ~ "in love"
  ))

```


One subsequent step could be to use the new variable in a $\chi^2$ test:

```{r}
xchisq.test(tip_gruppe ~ sex, data = tips)
```


Similarly, use `case_when` for nominal variables:

```{r}
tips <- tips %>% 
  mutate(weekend = case_when(
    day == "Fri" ~ "weekend",
    day == "Sat" ~ "weekend",
    TRUE ~ "keep on working"
  ))
```

Note that `TRUE` indicates "else do ...", in this case read "else 'weekend'  is 'keep on working'".

A convinient way to bin several values (such as "Fri", "Sat") into one (such as "weekend") is the `%in` operator:

```{r}
tips <- tips %>% 
  mutate(weekend = case_when(
    day %in% c("Fri", "Sat") ~ "weekend",
    TRUE ~ "keep on working"
  ))
```



Another convenient way is using `rec` from the r package `sjmisc`:

```{r}
library(sjmisc)
```


```{r}
tips <- rec(tips, day,
            rec = "Fri=Weekend; Sat=Weekend; else = keep_working")

count(tips, day_r)
```


Note that a new, recoded variable is appended using the suffix `_r`. See:

```{r}
glimpse(tips)
```


Note that the pipe will work too:

```{r}
tips <- tips %>% 
  rec(day,
      rec = "Fri=Weekend; Sat=Weekend; else = keep_working")

```

`rec` is convenient as one does not need to use `mutate`. 

Use `?rec` for more infos.


The good thing on both ways (`case_when` and `rec`) is that both functions can be used both for recoding and some binning purposes.

