---
title: Visualizing a regression plane (two predictors)
author: ''
date: '2018-12-13'
draft: FALSE
slug: visualizing-a-regression-plane-two-predictors
categories:
  - rstats
tags:
  - rstats
  - plotting
editor_options: 
  chunk_output_type: console
---



Plotting a "simple" regression (one regression) is pretty straight forward in R.


# Setup

```{r message = FALSE}
library(tidyverse)
data(mtcars)
library(mosaic)
library(modelr)
library(plotly)
```

# Define model


```{r}
lm1 <- lm(mpg ~ hp, data = mtcars)

mtcars <- mtcars %>% 
  mutate(lm1_pred = predict(lm1))
```


# Plot

One way:

```{r}
ggplot(mtcars) +
  aes(y = mpg, x = hp) +
  geom_point() +
  geom_lm()
```


Another way:

```{r}
ggplot(mtcars) +
  aes(x = hp) +
  geom_point(aes(y = mpg)) +
  geom_point(aes(y = lm1_pred), color = "blue") +
  geom_line(aes(y = lm1_pred), color = "blue")
```


Using  the `ggformula` interface to `ggplot2`:


```{r}
gf_point(mpg ~ hp, data = mtcars) %>% 
  gf_lm()
```



# 3D scatterplot with regression plane

## Define model with two predictors

```{r}
lm2 <- lm(mpg ~ hp  + wt, data = mtcars)

mtcars <- 
  mtcars %>% 
  mutate(lm2_pred = predict(lm2))
```



## Plot with plotly


Define prediction grid as the cartesian product of (the unique values of) weight and hp.

```{r}
wts <- unique(mtcars$wt)
hps <- unique(mtcars$hp)

wts <- seq(min(mtcars$wt), max(mtcars$wt), length.out = 25)
hps <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 25)

pred_grid <- expand.grid(wts, hps) %>% 
  rename(wt = Var1,
         hp = Var2) %>% 
  add_predictions(lm2)
```

Plotly needs the prediction data in matrix form.


```{r}
pred_mx <- matrix(pred_grid$pred,
                  ncol = 25,
                  nrow = 25)
```


And plot it:

```{r}
plot_ly() %>% add_surface(x = ~ 1:5, 
                          y = ~ 50:300, 
                          z = ~ pred_mx
                          )
```

