---
title: Compute all pairwise differences in matrix
author: ''
date: '2018-11-21'
categories:
  - rstats
tags:
  - rstats
slug: compute-all-pairwise-differences-in-matrix
editor_options: 
  chunk_output_type: console
---



A quite frequent task in many fields of applied math is to compute pairwise differences of elements in a matrix. Actually, it need not be a difference; a product is frequent, too. In this post, we explore some (base) R ways to achieve this. 


```{r message = FALSE}
library(mosaic)
library(gdata)
library(tidyverse)
```


# Using `outer()`


An elegant approach, using base R, is applying `outer()`. That's useful if one has two vectors, and wants to compute the outer product:

```{r}
v1 <- c(1:3)
v2 <- c(2:4)

outer(v1, v2)
```


As per default, `outer()` computes the product, because that's the definition of the outer product in linear algebra.

Notice that `v1` and `v2` are just simple numeric vectors (albeit named):
```{r}
str(v1)
```


However, the function is quite flexible; we can give some other function as well, say, a pairwise difference:

```{r}
o1 <- outer(v1, v2, FUN = "-")
o1 
```

What type of object is returned?

```{r}
class(o1)
str(o1)
```


Add names to the margins (ie., to the vectors):

```{r}
names(v1) <- v1
names(v2) <- v2
```

```{r}
o1 <- outer(v1, v2, FUN = "-")
o1
str(o1)
class(o1)
```


At times, we are interested rather in:

```{r}
vec <- c(v1, v2)
vec
str(vec)
o2 <- outer(vec, vec, FUN = "-")
str(o2)
o2
```

What we have just computed, is the difference between *all* pairs of elements, even between elements within the same vector. Previously, we have computed the difference between pairs of elements, where the first element is not part of the second vector (and v.v.).

If the absolute differences are needed, apply the `abs()` function on the whole matrix:

```{r}
o1_abs <- abs(o1)
o1_abs

o2_abs <- abs(o2)
```


If only the *different* combinations are needed (ie., without considering ordering), extract the upper triangle:

```{r}
o1_different <- upper.tri(o1) * o1_abs
o1_different

o2_different <- upper.tri(o2) * o2_abs
```


Notice that we do not want to use regular matrix multiplication, but cellwise computation only. For matrix multiplication, use `%*%`.


To get the *mean absolute pairwise difference*, take the mean of these values:

```{r}
mean(o1_different)
```

Wait; that's probably not what we expected. Let's tell R only to consider the values of the upper triangle when computing the mean (or whatever function, for that purpose).


```{r}
o1_diff_upptri <- ifelse(o1_different == 0, NA, o1_different)
o1_diff_upptri
```

More elegant, or, well, more direct, and much faster (no for-loops under the hood):


```{r}
o1_diff_upptri <- o1_different
o1_diff_upptri[o1_diff_upptri == 0] <- NA
o1_diff_upptri
```

That's a point where R shines: Not only vector arithmetic are applied automatically, also matrices can be digested.

Hang on, this one is nice, too:

```{r}
!o1_different
```

The `!` operator transform the matrix in a logical one (`0` = FALSE, else `TRUE`), and negates the values. 


Now:



```{r}
o1_variant <- o1_different
is.na(o1_variant) <- !o1_variant
o1_variant
```


However, those methods fail if there are zeros in the upper triangle:

```{r}
o2_variant <- o2_different
is.na(o2_variant) <- !o2_variant
o2_variant
```

We would *not* want the NAs in the upper triangle.

That's why this way comes handy:

```{r}
o2_diff_upptri <- o2_different
gdata::lowerTriangle(o2_diff_upptri, diag = TRUE) <- NA
o2_diff_upptri
```

Notice, however, that this function is not from base R.


Now, again the mean:

```{r}
mean(o1_diff_upptri, na.rm = TRUE)
sum(o1_diff_upptri, na.rm = TRUE)
```

And for `o2`:

```{r}
o2_mean <- mean(o2_diff_upptri, na.rm = TRUE)
o2_mean
```


Let's also compute the sum of the absolute differences:

```{r}
o2_sum <- sum(o2_diff_upptri, na.rm = TRUE)
o2_sum
```

What's the number of different elements?

```{r}
o2_length <- length(o2_diff_upptri[!is.na(o2_diff_upptri)])
o2_length
```

Divide sum by n to get the mean:

```{r}
o2_sum / o2_length
```


# Using `MAD()`

Incidentally, that's similar to what the functions `SAD()` and `MAD()` from `mosaic` compute. That is, the sum (`SAD()`) is the same:

```{r}
mosaic::SAD(v1, v2)
mosaic::SAD(v1, v2) == o2_sum
```

But the mean is different:

```{r}
mosaic::MAD(v1, v2)

mosaic::MAD(v1, v2)  == o2_mean
```


Internally, `MAD` does the following:

```{r}
x <- c(v1, v2)
x
M <- outer(x, x, "-")
M
M2 <- upper.tri(M) * abs(M)
M2
M3 <- sum(M2)
M3
length_elements <-  length(v1) + length(list(v2))
length_elements
M4 <- M3 / (length_elements)
M4
```

So, we divided by the number of the upper triangle elements. That is, by all pairwise comparisons. `MAD()` divides by the length of the first vector + 1. Not quite sure when we would want that. Presumably, these functions are not meant to be given two vectors, only one.

However, if we only have one vector, `MAD()` makes sense:

```{r}
MAD(vec)
```

Here, first the sum of the differences is computed:

```{r}
SAD(vec)
SAD(vec) == o2_sum
```

Next, this sum is divided by the number of elements (of `vec`):

```{r}
vec_length <- length(vec)

SAD(vec) / vec_length
SAD(vec) / vec_length == MAD(vec)
```



A glimpse to the code of those two functions:

```{r}
MAD_
```


```{r}
SAD_
```



# Debrief

We have explored some ways to compute the pairwise differences of two vectors, with some thoughts walking astray. We have discussed two ways of computing the "mean" by dividing by the number of comparisons or by the number of elements.