---
title: Visualizing mean values between two groups  - the tidyverse way
author: Sebastian Sauer
date: '2018-06-10'
slug: visualizing-summary-statistics-the-tidyverse-way
categories:
  - rstats
tags:
  - rstats
  - plotting
  - visualization
---

```{r echo = FALSE}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  0.4,  #0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)

```



A frequent job in data visualizing is to present summary statistics. In this post, I show one way to plot mean values between groups using the tidyverse approach in comparison to the mosaic way.

```{r}
library(tidyverse)
data(mtcars)
library(mosaic)
library(knitr)
library(sjmisc)
library(sjPlot)

```


# Visualizing mean values between two groups

First, let's compute the mean hp for automatic cars (`am == 0`) vs. manual cars (`am == 1`).
```{r}
mtcars %>% 
  group_by(am) %>% 
  summarise(hp_am = mean(hp)) -> hp_am
```

Now just hand over this data frame of summarized data to ggplot:

```{r}
hp_am %>% 
  ggplot() +
  aes(x = factor(am), y = hp_am) +
  geom_point(aes(shape = factor(am)), size = 5, alpha = .7) +
  geom_line(group = 1) -> p1
p1
```

An alternative approach would be using the R package `mosaic`:

```{r}
favstats(hp ~ am, data = mtcars) %>% 
  rownames_to_column() -> mtcars_favstats
```


Now hand over this data frame to `gf_XXX()` from `mosaic` (wrappers around `ggplot()`):

```{r}
gf_point(mean ~ rowname, data = mtcars_favstats, shape = ~rowname, size = 5, alpha = .7) %>% 
  gf_line(mean ~ rowname, group = 1)
```


# Adding uncertainty measures

If we are not primarily interested in sample description but population estimation we need to add uncertainty measures - such as the standard error (SE).

One easy solution is to make use of the package `gplots`:

```{r}
library(gplots)
plotmeans(hp ~ am, data = mtcars)
```

Note that the errors describe the 95% confidence interval of the respective mean.


Using ggplot2, there are some helper functions from `hmisc` that will do to computation of the standard error for us. One way is to use the `stat_summary()` function:

```{r}

p1 + 
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", data = mtcars, aes(y = hp))

```



`stat_summary()` provides an alternative to `geom_XXX()` for building a plot. Here the focus lies on "which summary statistic do I want to compute?

`mean_cl_normal` computes the *m*ean and the *c*onfidence *l*imits based on a t-distribution. As an alternative `mean_cl_boot` would produce a less assumption laden bootstrapped confidence interval.


`geom_errorbar()` yields essentially the same result. Here, we also specify that we do not want the 2-SE-interval (95%) but the 1-SE-interval (via `fun.args`).

```{r}
p1 +
  geom_errorbar(stat = "summary", fun.data = "mean_sdl", 
                  fun.args = list(mult = 1),
                  position =  position_dodge(width = 0.9),
                data = mtcars,
                aes(y = hp))
```


`stat = "summary"` computes a summary at each position of x. In this case, the summary function `mean_sdl` is called. `fun_data` returns three values: `y`, `ymin` and `ymax`.

# Less intrusive bars

I personally find the bars to intrusive; let's try to deemphasize their appearance.


```{r}

mtcars %>%
  ggplot() +
  aes(x = factor(am), y = hp) +
  geom_pointrange(stat = "summary", fun.data = "mean_sdl", 
                  fun.args = list(mult = 1),
                  position =  position_dodge(width = 0.9),
                size = 2,
                aes(y = hp, shape = factor(am), color = factor(am))) +
  geom_line(stat = "summary", fun.data = "mean_sdl",
            group = 1)
```


# Plot table

Sometimes, we do not (only) want a diagram but also the exact figures. How to get that? `kable()` provides a convenient way:

```{r}
hp_am %>% 
  kable()
```

The package `sjmisc` provides a convenience function for summary statistics:

```{r}
mtcars %>%
  group_by(am) %>% 
  select(hp) %>% 
  descr()
```


## Double grouping

Let's try a double grouping and see where it takes us:

```{r}
mtcars %>% 
  group_by(am, cyl) %>% 
  summarise(hp_mean = mean(hp)) %>%  
  ungroup() -> hp_am_cyl

hp_am_cyl
```

Now let's add the mean per `am` group (the "overall" means):

```{r}
hp_am_cyl %>% 
  group_by(am) %>% 
  mutate(overall_mean = mean(hp_mean)) -> hp_am_cyl

hp_am_cyl %>% kable()
```


This format is called the long format. We can shift it to the wide format:


```{r}
hp_am_cyl %>% 
  spread(key = cyl, value = hp_mean) %>% 
  kable()
```

