---
title: Showcase of Viridis, maps, and ggounty
author: Sebastian Sauer
date: '2018-05-18'
slug: showcase-of-viridis-maps-and-ggounty
categories:
  - dataviz
tags:
  - r
  - plotting
---


This posts shows how easy it can be to build an visually pleasing plot. We will use [hrbrmster](https://github.com/hrbrmstr)'s `ggcounty`, which is an R package at this [Github repo](https://github.com/hrbrmstr/ggcounty). Graphics engine is as mostly in my plots, Hadley Wickhams `ggplot`. All build on R. Standing on shoulders...

Disclaimer: This example heavily draws on hrbrmster example on [this page](https://github.com/hrbrmstr/ggcounty). All credit is due to Rudy, and those on whose work he built up on.




First, load the relevant packages:

```{r}
library(ggcounty)  # us maps
library(viridis)  # color scheme
library(tidyverse)  # data handling
```


Load population data:

```{r}
data(population, package = "ggcounty")
```


Split population figures in bins:


```{r}
population$brk <- cut(population$count, 
                      breaks=c(0, 100, 1000, 10000, 100000, 1000000, 10000000), 
                      labels=c("0-99", "100-1K", "1K-10K", "10K-100K", 
                               "100K-1M", "1M-10M"),
                      include.lowest=TRUE)
```


Get the US counties map:

```{r}
us <- ggcounty.us()
gg <- us$g
```

`us$g` contains base map:

```{r}
us$g
```


Add population data, this plot is called *choropleth*:

```{r}
gg <- gg + geom_map(data=population, map=us$map,
                    aes(map_id=FIPS, fill=brk), 
                    color="white", size=0.125)
```


Now give it the viridis color scheme:


```{r}
gg + scale_fill_viridis_d()
```


Besides the default scale of virids, there are some other scales, such as `magma` and `inferno`:



```{r}
gg + scale_fill_viridis_d(option = "magma") +
  theme(legend.position = "none")
```


Stunning, isn't it? And simple. Enjoy!


