---
title: Easy way to convert factors zu numbers
author: Sebastian Sauer
date: '2018-06-22'
slug: easy-way-to-convert-factors-zu-numbers
categories:
  - rsta
tags:
  - tutorial
  - rstats
---


```{r setup, echo = FALSE}
set.seed(1014)
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  0.4,  #0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)


options(max.print = 20,
        dplyr.print_max = 7,
        dplyr.print_min = 3)

```


Converting factors to numbers in R can be frustrating. Consider the following sitation: We have some data, and try to convert a factor (`sex` in `tips`, see below) to a numeric variable:

```{r}
library(tidyverse)
library(sjmisc)   # for recoding
data(tips, package = "reshape2")
glimpse(tips)
```

Ok, here we go. Say, we want `"Female" = 1` and `"Male" = 0`. 

```{r}
tips <- sjmisc::rec(tips, sex, rec = "Female = 1; Male = 0")
glimpse(tips)
```

Seems to have worked so far. Wait, `sex_r` ist still a factor, not numeric. So convert it using `as.numeric`:

```{r}
tips$sex_num <- as.numeric(tips$sex_r)
glimpse(tips)
```

Oh no! That's not what we wanted! R has messed the thing up (?). The reason is that R sees the first factor level internally as the number 1 . The second level as number two. What's the first factor level in our case? Let's see:

```{r}
factor(tips$sex) %>% head()
factor(tips$sex_r) %>% head()
```

That's confusing: "0" is the first level of `sex_r` - internally for R represented by "1". The second level of `sex_r` is "1" - internally represented by "2". That's why we get these numbers:

```{r}
head(tips$sex_num)
```


## Solution

One solution is to use `readr::parse_number()`:

```{r}
tips$sex_num <- parse_number(tips$sex_r)
head(tips$sex_num)
head(tips$sex_r)
```


Worked!