---
title: Map students to presentation slots
author: Sebastian Sauer
date: '2018-03-11'
slug: map-students-to-presentation-slots
categories:
  - rstats
tags:
  - rstats
  - tutorial
---

As a teacher, I not only teach but also assess the achievements of students. One example of a typical student assignments is a presentation. You know, powerpoint slides and stuff.

For that purpose, I often need to map students to one of several time slots. Here's the R code I use for that purpose.


```{r}
library(tidyverse)
```

How many students are subscribed to the assignment?

```{r}
stud_count <- 20
```

Let's say there `r stud_count` students in the course.


Let's assume, for each time slot, 10 students can be allocated (not more time for more student for one slot).

```{r}
slots_count <- (stud_count / 10) %>% ceiling  # round to next integer
slots_count
```

That gives `r slots_count` time slots.


Now let's map the students to the slots on a random base:


```{r}
slot <- 1:max(slots_count)

set.seed(2018)
allocation <- replicate(n = 10,
              sample(x = slot)) %>% as.vector

```


What we've done here is:

1. Take the first triplet of students and randomly assign them to the three slots (so that each of the three students has a unique slot, ie sampling without replacement)
2. Repeat that for the rest of the triplets



Let's check whether the `allocation` has worked:


```{r}
table(allocation)
```


Worked out. Presentations ahead.

Let's save the vector as a csv-file for the easy of interfacing with other applications such as Excel.


```{r}
library(rio)
export(data_frame(slots = allocation),
       file = "slots.csv")
```









