---
title: 'Playing around with geo mapping: combining demographic data with spatial data'
author: Sebastian Sauer
date: '2018-05-28'
csl: "/Users/sebastiansaueruser/Documents/Literatur/ref_mgt/apa.csl"
bibliography: "/Users/sebastiansaueruser/Documents/Literatur/ref_mgt/library.bib"
lang: de-De
slug: playing-around-with-geo-mapping-combining-demographic-data-with-spatial-data
categories:
  - rstats
tags:
  - rstats
  - plotting
  - visualization
  - geo
---


```{r echo = FALSE}
knitr::opts_chunk$set(message = FALSE, fig.align = "center")
```


In this post, we will play around with some basic geo mapping. More preciseyl, we will explore some easy ways to plot a choropleth map.

First, let's load some geo data from [Bundeswahlleiter](https://www.bundeswahlleiter.de/bundestagswahlen/2017/ergebnisse.html), and combine it with some socio demographic data from the same source.

## Preparation


Let's load some packages:

```{r}
library(tidyverse)
library(sf)
library(viridis)
suppressPackageStartupMessages(library(googleVis))
```

Geo data:
```{r}
my_path_wahlkreise <- "~/Documents/datasets/geo_maps/btw17_geometrie_wahlkreise_shp/Geometrie_Wahlkreise_19DBT.shp"
file.exists(my_path_wahlkreise)
```

socio demographic data:

```{r}
unemp_file <- "~/Documents/datasets/Strukturdaten_De/btw17_Strukturdaten-utf8.csv"
file.exists(unemp_file)
unemp_de_raw <- read_delim(unemp_file,
                           ";", escape_double = FALSE,
                           locale = locale(decimal_mark = ",",
                                           grouping_mark = "."),
                           trim_ws = TRUE,
                           skip = 8)
```




```{r}
wahlkreise_shp <- st_read(my_path_wahlkreise)
```



Some data cleansing:

```{r}
unemp_names <- names(unemp_de_raw)
unemp_names
unemp_de <- unemp_de_raw
names(unemp_de) <- paste0("V",1:ncol(unemp_de))

unemp_de <- unemp_de %>%
  rename(state = V1,
         area_nr = V2,
         area_name = V3,
         for_prop = V8,
         pop_move = V11,
         pop_migr_background = V19,
         income = V26,
         unemp = V47)  # total, as to March 2017
```


Join the two data files:

```{r}
wahlkreise_shp %>%
  left_join(unemp_de, by = c("WKR_NR" = "area_nr"))  -> choro_data
```



## Some choropleth maps

Some choropleth maps, such as of

... German income as per administrative area:

```{r}
choro_data %>%
  ggplot +
  geom_sf(aes(fill = income)) +
  theme_void() +
  scale_fill_viridis()
  
```



... unemployment rate with legend suppressed:


```{r}
choro_data %>%
  ggplot +
  geom_sf(aes(fill = unemp)) +
  theme_void() +
  scale_fill_viridis()  +
  theme(legend.position = "none")
```


... Proportion of foreigners with legend name set:

```{r}
choro_data %>%
  ggplot +
  geom_sf(aes(fill = for_prop)) +
  theme_void() +
  scale_fill_viridis() +
  labs(fill = "Foreigners") +
  theme(legend.direction = "horizontal",
        legend.position = "bottom")
```


Let's get some statistics out of the data:


```{r}
choro_data %>% 
  select(area_name, for_prop) %>% 
  top_n(for_prop, n =3) %>% 
  ggplot(aes(x = reorder(area_name, for_prop), y = for_prop, fill = for_prop)) +
  geom_col() +
 # scale_fill_viridis_c() +
  coord_flip()
  
```




```{r include = FALSE}
choro_data %>% 
  select(area_name, for_prop) %>% 
  top_n(for_prop, n =3) %>% 
  ggplot(aes(x = reorder(area_name, for_prop), y = for_prop)) +
  geom_col() +
 # scale_fill_viridis_c() +
  coord_flip() +
  labs(x = "Area",
       y = "Opportunity Strength") +
  scale_y_continuous(breaks = NULL) +
  scale_x_discrete(breaks = NULL) +
  geom_label(aes(x = area_name,
                y = 1,
                label = area_name),
            hjust = 0,
            size = 12)
  
```

