---
title: 'New split-apply-combine variant in dplyr: group_split()'
author: ''
date: '2018-12-10'
slug: new-split-apply-combine-variant-in-dplyr-group-split
categories:
  - rstats
tags: []
editor_options: 
  chunk_output_type: console
---


---

UPDATE 2018-12-11 - I'm talking about the package DPLYR, not PURRR, as I had mistakenly written.


---




There are many approaches for what is called the "split-apply-combine" approach (see [this paper](https://www.jstatsoft.org/article/view/v040i01/v40i01.pdf) by Hadley Wickham).


I recently thought about the best approach to use split-apply-combine approaches in R (see [tweet](https://twitter.com/sauer_sebastian/status/1071511342220222465), and this [post](https://data-se.netlify.com/2018/12/05/plot-many-ggplot-diagrams-using-nest-and-map/)).


And I retweeted some criticism on the "present era" tidyverse approach (see [this tweet](https://twitter.com/sauer_sebastian/status/1071511342220222465)), and check out the mentioned [post](https://coolbutuseless.bitbucket.io/2018/03/03/split-apply-combine-my-search-for-a-replacement-for-group_by---do/) by @coolbutuseless.



```{r echo = FALSE}
blogdown::shortcode('tweet', "1071511342220222465")
```



Then, Erich Neuwirth (`@neuwirthe`) informed me on Twitter, that there's a new idiom in ~~purrr()~~ dplyr^[Thanks @romain_francois for pointing out] (as of version 0.8.0) that might come as a remedy: `split_group()`. This post explore some of the uses of this idiom.


```{r echo = FALSE}
blogdown::shortcode('tweet', "1071538348387377152")
```


---

Thanks, Erich!

`r emo::ji("thumbsup")`

---




First, load the tidyverse packages (that is, dplyr, for our purposes):

```{r}
library(tidyverse)
library(broom)
```

Mind that I run this version of dplyr (loaded from github on 2018-12-10):

```{r}
packageVersion("dplyr")
```



# The typical tidyverse approach for split-apply-combine


The typical tidyverse approach is the following:
```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(hp_mean = mean(hp))
```

Contrast to the overall mean:

```{r}
mtcars %>% 
  summarise(mean(hp))
```


# Using `list()` in `summarise()` does NOT convey the grouping

This approach works fine unless one wants to apply a more complex function to each group. "Complex" refers to a function that yields back more than one number, such as `lm()`.

One might hope that the following works:


```{r}
d <- mtcars %>% 
  group_by(cyl) %>% 
  summarise(hp_mean = list(tidy(lm(hp ~ 1, data  = .))))

d
```

This gives what is called a "list column" in tidyverse parlance. To "unpack" or "unnest" this list column use `unnest()`. Notice that this only works if the list column is "tidy", that is, if it can be unpacked to a data frame like structure.


```{r}
d %>% 
  unnest()
```


However, the summarize function did apparently **not** consider the grouping.



# Using `split()` and `map()`


A more "purrr-ish" approach is this:

```{r}
mtcars %>% 
  split(.$cyl) %>% 
  map( ~ tidy(lm(hp ~ 1, data = .))) %>% 
  map_dfc("estimate")
```


Works. But.

This approach is different to the dplyr approach, mainly because the grouping idiom of dplyr does not work here (and `split(.$cyl)` looks not very consistent to the tidyverse code).



# Using `group_by()` and `nest()`ed list-columns


List columns do work when used with `group_by()` and `mutate()`:


```{r}

apply_lm <- function(df){
      tidy(lm(data = df, hp ~ 1))      
}

d <- mtcars %>% 
  group_by(cyl) %>% 
  nest() %>% 
  mutate(hp_mean = purrr::map(.f = apply_lm, . = data)) 
d
```


Let's see what's in the list columns:

```{r}
d %>% 
  unnest(hp_mean)
```



# Using `group_split()`


Since recently dplyr features `group_split()` (still experimental as of this writing).

The idea of this function is to convey the grouping information from `group_by()` to `split()`.

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  group_split() %>% 
  map_dfr(~lm(hp ~ 1, data = .) %>% tidy())
```

When the data frame is ungrouped, `group_split()` can be used to group it:

```{r}
mtcars %>% 
  group_split(cyl) %>% 
  map_dfr(~lm(hp ~ 1, data = .) %>% tidy())
```


# Join model results back to full (raw data) data frame

This dataset could now joined with the initial data:


```{r}
mtcars %>% 
  group_split(cyl) %>% 
  map_dfr(~lm(hp ~ 1, data = .) %>% tidy()) %>%
  mutate(group = mtcars %>% group_keys(cyl) %>% pull(cyl)) %>% 
  full_join(mtcars, by = c("group" = "cyl"))
```

Of course, the join builds a lot of duplicated data which can be a waste.



# One wish remains open  - Adding columns using `group_slit()`


However, at the moment at least, I do not see a convenient way to add columns to a existing data frame (e.g., add the estimate) via this approach:


```{r error = TRUE}
mtcars %>%
  mutate(hp_mean = group_split(., cyl) %>% map(~lm(hp ~ 1, data = .) %>% tidy() %>% map("estimate")))
  
```




The error tells us that the nested version seems more reasonable:


```{r}
mtcars %>% 
  group_by(cyl) %>%
  nest() %>% 
  mutate(hp_mean = map(.f = apply_lm, .x = data)) %>% 
  unnest(hp_mean)
```

Individual columns can also be extracted using `map()` followed by the name of the element:

```{r}
mtcars %>% 
  group_by(cyl) %>%
  nest() %>% 
  mutate(hp_mean = map(apply_lm, .x = data)) %>% 
  mutate(hp_est = map_dbl("estimate", .x = hp_mean))
```


# The same without externally defined function

It may be more direct to define the function for `map()` right within `mutate()`:

```{r}
mtcars %>% 
  group_by(cyl) %>%
  nest() %>% 
  mutate(hp_lm = map( ~ lm(hp ~ 1, data = .) %>% tidy(), .x = data)) %>% 
  mutate(hp_mean = map_dbl("estimate", .x = hp_lm))
```



# Debrief


`split_groups()` is a new function (as of today) which provides some remedy for split-apply-combine actions within the tidyverse. 
