---
title: Quick intro to geo plotting
author: Sebastian Sauer
date: '2018-05-17'
slug: quick-intro-to-geo-plotting
categories:
  - rstats
  - dataviz
tags:
  - r
  - plotting
---

```{r echo = FALSE}
set.seed(1014)
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  0.4,  #0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)


options(max.print = 20,
        dplyr.print_max = 7,
        dplyr.print_min = 3)

```


Geo plotting can be simple; at least in its basic form. Let's review an example using data from [GADM](https://gadm.org/index.html).

GADM provides maps for many (all?) countries of the world; not only the country borders but even administrative border lines an a finder scale.

It comes very handy that the map (geo) can be downloaded as R data files:

```{r}
data_path <- "https://biogeo.ucdavis.edu/data/gadm3.6/Rsf/gadm36_DEU_2_sf.rds"
```

Load some packages:


```{r}
library(sf)
library(tidyverse)
library(downloader)
```


Download the maps data:

```{r}
download(url = data_path, "d.Rds")
d <- readRDS("d.Rds")
```


We have downloaded map data from Germany, detail level 2 (administrative discticts "Landkreise"). Now plot:

```{r out.width="100%", out.width=5}
d %>% 
  ggplot() + 
  geom_sf()
```


Data for each Landkreis can be access from various sources, most prominently maybe, for Germany, the Statistisches Bundesamt.

Here are some unemployment figures from this source:

```{r}
download_osf <- function(repo){
  osf_link <- paste0("https://osf.io/",repo, "/?action=download")
  out <- read.csv(osf_link)
  return(out)
}

alz <- download_osf("mrxh3")
```

We can now try to join the two tables, but it turns out this may not be so easy as the district name are only similar, not identical.

Let's do some checks.

```{r}
library(stringr)
str_detect(d$NAME_1, as.character(alz$name) )
```

