---
title: Metrics for readability of slide decks for teaching
author: Sebastian Sauer
date: '2018-04-09'
slug: metrics-for-readability-of-slide-decks-for-teaching
draft: TRUE
categories:
  - rstats
tags:
  - textmining
  - r
---


Some thoughts on the "quality" of slide decks used in teaching: How many visuals are there? How much text is squeezed on a slide? What's the average word and sentence length? How much buzz words and complicated jargon is present?


Let's have an initial look at some meaures in that regard. Hey, this post is just playing around a little. We will work with this [slide deck](https://github.com/sebastiansauer/sebastiansauer.github.io/blob/master/download/slides.pdf), published under CC-BY 3, with parts as GNU - so all "open" in short. That deck comprises teaching material we have compiled for stats courses at our institution; Karsten Luebke at teh driver seat. Norman Markgraf did the TeX magic and more.


Let's parse the PDF file first:


```{r message = FALSE}
library(pdftools)
library(tidytext)
library(tidyverse)
library(stringr)
library(mosaic)
library(stringi)

file_url <- "https://github.com/sebastiansauer/sebastiansauer.github.io/raw/master/download/slides.pdf"
download.file(file_url, "slidedeck.pdf")
slidedeck <- pdf_text(file_url)
```


How does that text object look like:

```{r}
head(slidedeck)
```


Let's have a look at the distribution of word count per page:


```{r}
slidedeck %>% 
  map_int(~str_count(.)) %>% 
  as_tibble() -> slide_word_count

slide_word_count %>% 
  ggplot() +
  aes(x = value) +
  geom_histogram()
```


Some quantiles:

```{r}
mosaic::qdata(~value, data = slide_word_count)
```

