---
title: Finding NAs in multiples columns (per row)
author: ''
date: '2018-08-09'
slug: finding-nas-in-multiples-columns-per-rows
categories:
  - rstats
tags:
  - rstats
---


Assume you would like to check for missing data, but not for one column only but for several columns. 

First, data and some packages:

```{r}
data(mtcars)
library(tidyverse)
```

Then, let's introduce some missing data:

```{r}
mtcars[c(1,2), 1] <- NA
mtcars[c(1, 3:4), 2] <- NA
```


# Don't check columns individually

Of course, you do not want to repeat yourself, and check each column individually, like this:

```{r}
sum(is.na(mtcars[[1]]))
sum(is.na(mtcars[, 1]))  # same
```

Neither one would like to check each row individually:

```{r}
sum(is.na(mtcars[1, ]))
```


# Apply a function to each column

We need to `apply()` the function above to each column (or row). `map` works similar to `apply` but has some niceties included (`map` comes from R package `purrr`).


```{r}
mtcars %>% 
  map(~sum(is.na(.)))
```


Note that `~` is shorthand for `function(x)`, but less verbose. `.` refers to each element of `mtcars`, ie each column.


# Count Nas per row

Now assume you want to know the missing per case (ie row). One way is this:

```{r}
mtcars %>% 
mutate(NA_count = rowSums(is.na(.))) %>% 
  head()
```

Here, the dot `.` refers to the data frame as of the last pipe step. In this case, it's just the plain data frame `mtcars`.


Note that `is.na()` expects a data.frame as input.

```{r}
is.na(mtcars) %>% 
  head()
```

Note that `sum()` also accepts a data.frame as input:

```{r}
is.na(mtcars) %>% 
  sum() 
```


# Some musings

To inspect missing values (or cases with NAs), `filter()` may be a solution:


```{r}
mtcars %>% 
  filter(is.na(mpg) |  is.na(cyl) | is.na(disp))
```


More conveniently using `complete.cases()`:


```{r}
complete.cases(mtcars)
```

And the complement:


```{r}
mtcars %>% 
  filter(!complete.cases(mtcars))
```

