---
title: What are the names of the cars with 4 cylinders?
author: ''
date: '2018-12-03'
slug: what-are-the-names-of-the-cars-with-4-cylinders
categories:
  - rstats
tags:
  - rstats
  - tidyverse
editor_options: 
  chunk_output_type: console
---

```{r knitr-steup, echo = FALSE}

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "100%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =   .618,
  fig.show = "hold",
  size = "tiny"
)
```




Recently, some one asked me in a workshop this question: "What are the names of the cars with 4 (6,8) cylinders?" (he referred to the `mtcars` data set). That was a workshop on the tidyverse, so the question is how to answer this question using tidyverse techniques.

First, let's load the usual culprits.

```{r}
library(tidyverse)
library(purrrlyr)
library(knitr)
library(stringr)
```



```{r}
data(mtcars)

d <- as_tibble(mtcars) %>% 
  rownames_to_column(var = "car_names")

d %>% 
  head() %>% 
  kable()
```


Let's explore several ways.

# Way 1 - using `paste()` aund `pull()`

First, it makes sense to group the data, as our question refers to the car names of each cylinder group. Next, we can summarize the data using `paste()` to collate the names into one string:

```{r}
d2 <- 
  d %>% 
  group_by(cyl) %>% 
  summarise(car_names_coll = paste(car_names, collapse = " "),
            hp_mean_per_cyl = mean(hp))

d2

d2 %>% 
  pull(car_names_coll)
```




# Way 2 - using `nest()`

Of interest, list columns provide a more data frame type answer. We "fold" or "nest" a vector, list, or data frame in one cell:

```{r}
d %>% 
  group_by(cyl) %>% 
  nest(car_names)
```



How to access it?

```{r}
d_nest <- d %>% 
  group_by(cyl) %>% 
  nest(car_names)


d_nest$data[[1]]
d_nest[[1, "data"]]
```



How to `unnest`?

Simple:

```{r}
d_nest %>% 
  unnest()
```


# Difference between `[` and `[[`


By the way, that's the difference between `[` and `[[`? Or is there none?

```{r}
identical(d_nest$data[[1]] , d_nest$data[1])
```


There is. See the difference between the `class`es:


```{r}
d_nest$data[[1]] %>% class()
d_nest$data[1] %>% class()

```


`[[` extracts the actual object, a data frame.

`[` extracts a list (containing the actual object).


# Processing list columns


How to process list column data further?

Say we would like to know if there ist least one Mercedes in each group:


```{r}
d_nest %>% 
  mutate(mercs_lgl = str_detect(data, "Merc"))
```

We get a warning because `data` is not vector, but a data frame (with one column, ie., the car names).

And how many Mercs are there in each group?


To start, consider this way of counting instances of Mercs:

```{r}
str_detect(d_nest$data[[1]]$car_names, "Merc") %>% sum()
```


Next, we build that into our dplyr pipeline:


```{r}
d_nest %>% 
  mutate(names_list = map(data, "car_names")) %>% 
  mutate(mercs_n = map(names_list, ~{str_detect(., pattern = "Merc") %>% sum()})) %>% 
  unnest(mercs_n)
```


Phish, that looks frightening. Let's go throught it:

- First, we pull out the names of the cars, because `str_detect()` is not happy to work with a data frame as input. Now we have the column `names_list` which is of type character
- Second, we map `str_detect` to each row of this new column `names_list`.  This fuctions looks for Merc instances. Wait, we want to sum up the Marc instances, that's way we also use `sum()`. The curly braces make sure that only the last evaluated expression is handed back.
- Third, we have to unnest the `mercs_n` column, because it is still in nested format even it consists of one value only.






# Way 3 - like using `nest` but without a special idiom

Again, we fall back to the classical dplyr way of summarising data to a single (scalar) value:

```{r}
d3 <-
  d %>% 
  group_by(cyl) %>% 
  summarise(names_per_cyl = list(car_names),
            hp_mean_per_cyl = mean(hp))

d3$names_per_cyl[[1]]
```



# Fold 'em all in a list data frame

Let's take the nesting into list data frames to its extreme:


```{r}
d %>% 
  group_by(cyl) %>% 
  summarise_all("list")
```




# Way 4 - using `map()`:



```{r}
d %>% 
  split(.$cyl) %>% 
  map("car_names")
```



# Debrief


`summarise()` in dplyr, summarizes a (column) vector to a scalar (single value). This is often handy, but sometimes limiting, as you are only allowed to apply functions that return a scalar. For more complex functions, such as `lm()` different approaches need be chosen. One way is to work with list columns as they provide a neat way to plug stuff into one cell of a data frame. More flexible approaches can be built upon the family of `apply` or `map()` and alike.


