---
title: Visualizing a multivariate normal distribution
author: ''
date: '2018-12-13'
slug: visualizing-a-multivariate-normal-distribution
draft: FALSE
categories:
  - rstats
tags:
  - plotting
editor_options: 
  chunk_output_type: console
---


In R, it is quite straight forward to plot a normal distribution, eg., using the package `ggplot2` or plotly.

# Setup

```{r message = FALSE}
library(tidyverse)
library(mvtnorm)
library(plotly)
library(MASS)
```




# Simulate multivariate normal data

First, let's define a covariance matrix $\Sigma$:

```{r}
sigma <- matrix(c(4,2,2,3), ncol = 2)
sigma
```

Then, simulate observations n = `n` from these covariance matrix; the means need be defined, too. As the rank of our covariance matrix is 2, we need two means:

```{r}
means <- c(0, 0)
n <- 1000
```



```{r}
set.seed(42)
x <- rmvnorm(n = n, mean = means, sigma = sigma)
```


What's the structure of `x`?

```{r}
str(x)
```

Let's make a data frame out of it:

```{r}
d <- data.frame(x)
names(d)
```


Now, let's plot.

# Plotting

## Plotting univariate (sampled) normal data

Well, that's obvious.

```{r}
d %>% 
  ggplot(aes(x = X1)) +
  geom_density()
```


### Plot theoretic normal curve


```{r}
p1 <- data_frame(x = -3:3) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = dnorm, n = n)

p1
```


### Interactive version of normal curve

```{r}
ggplotly(p1)
```

## Plotting multivariate data


### 2D density

```{r}
p2 <- ggplot(d, aes(x = X1, y = X2)) +
  geom_point(alpha = .5) +
  geom_density_2d()

p2
```

### Same with plotly

```{r}
ggplotly(p2)
```


### Geom binhex


```{r}
p3 <- ggplot(d, aes(x = X1, y = X2)) +
  geom_point(alpha = .5) +
  geom_bin2d() +
  scale_fill_viridis_c()

p3
```


### 2D histogram (heatmap) with plotly

```{r}
(p <- plot_ly(d, x = ~X1, y = ~X2))

add_histogram2d(p)

```

### 2D cntour with plotly

```{r}
add_histogram2dcontour(p)
```


## 3D plot

#### Surface

```{r}
dens <- kde2d(d$X1, d$X2)

plot_ly(x = dens$x,
        y = dens$y,
        z = dens$z) %>% add_surface()
```

### 3D Scatter


First, compute the density of each (X1, X2) pair.

```{r}
d$dens <- dmvnorm(x = d)
```


Now plot a point for each (X1, X2, dens) tuple.


```{r}
p4 <- plot_ly(d, x = ~ X1, y = ~ X2, z = ~ dens,
              marker = list(color = ~ dens,
                            showscale = TRUE)) %>% 
  add_markers()

p4
```

