---
title: How to create columns in a dataframe in R
author: Sebastian Sauer
date: '2018-03-07'
slug: how-to-create-columns-in-a-dataframe-in-r
categories:
  - rstats
tags:
  - rstats
  - tutorial
---


Note that we will use this library for this post:
```{r}
library(dplyr)
```


By the way, loading `mosaic`, will load `dplyr` too.

One of the major data wrangling activities (in R and elsewhere) is to create a new column in a data frame. For example, assume you have some students who have completed some exercises. In each *row* of the dataframe - one student. In each column - one exercise (called item). The dataframe might then look like this:



```{r}

df <- data.frame(name = c("John", "Joan", "Jeanne"),
                 item1 = c(TRUE, TRUE, FALSE),
                 item2 = c(TRUE, FALSE, TRUE))
df
```


Note that `TRUE` indicates that the exercise has been solved correctly, if otherwise, we note `FALSE`. 

Now assume you would like to sum up the number of correct items: How many items were solved by John, Joan, and Jeanne, respectively?


R provides a number of ways to compute this.

# Sum as a vector

```{r}
df$item1 + df$item2
```


So, John, solved 2 items, the other students 1 item. It comes handsome that R interpretes `TRUE` as 1 and `FALSE` as 0. As a consequence, we can add up logical variables with no hazzle.

But we don't want that as a freefloating column. That sum-column should be attached to our data frame. There should be a `items_sum` column in our dataframe.

# Base R way to create a column

The Base R way is this:


```{r}
df$items_sum <- df$item1 + df$item2
```

Let's check the dataframe:

```{r}
df
```

Worked. But there are other ways too.


# Using `dplyr`


```{r}
df <- mutate(df, items_sum2 = item1 + item2)
df
```

Worked. Now let's use the pipe.


# Using the pipe


```{r}
df <- df %>% 
  mutate(items_sum3 = item1 + item2) 
df
```


Note that the assignment arrow operator can be headed to either direction, ie left or right:


```{r eval = TRUE}
df %>% 
  mutate(items_sum3 = item1 + item2) -> df
df
```


# Recode variable with `car::Recode`


In R parlance, recoding a variable is different from creating one. Again, different ways exist for recoding variable. Let's assume that we would like to recode `TRUE` to 1 and `FALSE` to 0. But if we want to save the recoded variable as a new variable, then yes, we need to create a new variable with the same machinery as above. For that purpose, the package `car` comes handy:

```{r}
library(car)
df %>% 
  mutate(item1r = car::Recode(item1, 
                              "1 = TRUE; 0 = FALSE")) -> df
df

```





`car::recode` can be used for more complex recoding schemes, too:

```{r}
df <- df %>% 
  mutate(items_sum3 = car::recode(items_sum, 
                                 "lo:1 = 'failed';
                                  2 = 'passed';
                                  3:hi = 'cant believe'")) 
df
```

See `?Recode` for details of the syntax.


# Recode variable with `ifelse`


Another frequently used way is using `ifelse`:

```{r}
df %>% 
  mutate(item2r = ifelse(item2 == TRUE, 1L, 0L)) -> df
df
```

The general form of `ifelse` is `ifelse(condition, what_if_true, what_if_not)`.

By the way, if we want to coerce a numeric variable to integer, we can call it `1L`, instead of `1`.


# Recode using `case_when`


The more general form of `ifelse` is `case_when`. If generalizes `ifelse` to more than 2 cases:


```{r}
df %>% 
  mutate(items_sum4 = case_when(
    items_sum < 2 ~ "failed",
    items_sum == 2 ~ "passed",
    items_sum > 2 ~ "awesome")) -> df
df
```

