---
title: Correlation cannot be more extreme than +1/-1, proof using Cauchy-Schwarz
  inequality
author: Sebastian Sauer
date: '2019-11-19'
slug: correlation-cannot-be-more-extreme-than-1-1-proof-using-cauchy-schwartz-inequality
categories:
  - rstats
tags:
  - correlation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```



# The correlation coefficient cannot exceed an absolute value of 1

This is well-known. But why is that the case? How can we proof it? This post gives one explanation using the Cauchy-Schwarz inequality.


Here's one version of the definition of correlation:


$$
r = \frac{\sum(\Delta x \Delta y)}{\sqrt{\sum \Delta x^2} \sqrt{\sum \Delta y^2}}
$$



where $\Delta x$ and $\Delta y$ are the *differences* of $x_i$ and $\bar{x}$, that is: $\Delta x_i = x_i - \bar{x}$, and similarly for $\Delta y_i$.

For the ease of notation, let's proceed with the understanding that $x$ stands for the *differences*, ie $\Delta x$ (and similarly for $y$):



$$
r = \frac{\sum(xy)}{\sqrt{\sum x^2} \sqrt{\sum y^2}}
$$


Now, we conjecture that


$$
r = \frac{\sum(xy)}{\sqrt{\sum x^2} \sqrt{\sum y^2}} \le 1
$$

Let's multiply the equation by the denominator of the LHS:


$$
\sum(xy) \le \sqrt{\sum x^2} \cdot  \sqrt{\sum y^2}
$$


The [Cauchy Schwarz inequality](https://en.wikipedia.org/wiki/Cauchy%E2%80%93Schwarz_inequality) states that


$$
\big| \langle x,y\rangle \big |\leq ||x|| \cdot ||y||
$$

In words, the [inner product](https://en.wikipedia.org/wiki/Inner_product_space) $\langle x,y\rangle$ (in its positive variant, ie $>0$) is smaller or equal to the product of the [vector norms](http://mathworld.wolfram.com/VectorNorm.html).


Stated differently:

$$
\sum xy \le \sqrt{\sum x^2} \cdot  \sqrt{\sum y^2}
$$

Which is what we wanted to proof in the first place.


Here's a quite [nice intuition](https://jakubmarian.com/intuition-behind-the-cauchy-schwarz-markovs-and-jensens-inequality/) on the Cauchy Schwarz inequality.

