---
title: How to mutate all columns of a data frame
author: Sebastian Sauer
date: '2019-03-13'
slug: how-to-mutate-all-columns-of-a-data-frame
categories:
  - rstats
tags:
  - tidyverse
editor_options: 
  chunk_output_type: console
---



Say, you have a data frame with a number of columns, and you need to change every column in a similar way. A common example might be to standardize all (numeric) variables. How to do that in R? This post shows and explains an example using `mutate_all()` from the tidyverse.

Let's stick to the question "how to z-standardize all columns" for the sake of simplicity (and neglect that there are precooked solutions, for example from the superb package `sjmisc` by strengejacke.)


```{r message=FALSE}
library(tidyverse)
data(iris)
```



# Easy but inefficient way

```{r}
iris %>% 
  mutate(Sepal.Length_z = (Sepal.Length - mean(iris$Sepal.Length)) / sd(iris$Sepal.Length),
         Sepal.Width_z = (Sepal.Width - mean(iris$Sepal.Width)) / sd(iris$Sepal.Width))  %>% 
  head()
```

Beware the braces; it's easy to get bitten (happened to me).

Clearly, this appraoch dow not scale well. In addition, you'll strain your hand. And will enjoy the funniest typos.



# Define helper function

A first useful step is to define a helper function which we will apply on every column:

```{r}
z_std <- function(observed) {
  
  result <- (observed - mean(observed)) / sd(observed)
}
```

Of course, such a fucntion already exists a myriad times in other scripts, and yes, it is not crafted beautifully, but it will serve as a prgramatic start.

Now let's apply it:

```{r}
iris %>% 
  mutate(Sepal.Length_z = z_std(Sepal.Length)) %>% 
  head()
```


Much cleaner, simpler, more relaxing.



# Now to the conveyer belt

Now let's apply it to each column:


```{r}
iris %>% 
  select_if(is.numeric) %>% 
  mutate_all(funs(z = z_std(.))) %>% 
  head()
```



# Changes in dplyr


You might have noticed this warning:

`Warning: funs() is soft deprecated as of dplyr 0.8.0`


So let's change the code above to reflect the change in dplyr.


```{r}
iris %>% 
  select_if(is.numeric) %>% 
  mutate_all(list(z = ~ z_std(.))) %>% 
  head()
```


This code can be written more compact, see below.


```{r}
iris %>% 
  select_if(is.numeric) %>% 
  mutate_all( ~ z_std(.)) %>% 
  head()
```


Note that if you *don't* supply a name (suffix) such as `z` in the example above, the function will silently overwrite the original variables.

