---
title: Generalized rowwise operations using purrr::pmap
author: Sebastian Sauer
date: '2019-07-03'
slug: generalized-rowwise-operations-using-purrr-pmap
categories:
  - rstats
tags:
  - rowwise
  - purrr
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```


Rowwwise operations are a quite frequent operations in data analysis. The R language environment is particularly strong in *column* wise operations. This is due to technical reasons, as data frames are internally built as column-by-column structures, hence column wise operations are simple, rowwise more difficult.

This post looks at some rather general way to comput rowwise statistics. Of course, numerous ways exist and there are quite a few tutorials around, notably by [Jenny Bryant](https://github.com/jennybc/row-oriented-workflows), and by [Emil Hvitfeldt](https://www.hvitfeldt.me/blog/purrr-tips-and-tricks/) to name a few.


Let's work with these data:

```{r}
data(iris)
```



# Example 1

Assume you'd like to compute a a row sum for each case. Let's neglect that there exist a function called `rowSums` which does this job nicely. However, there might be situations where the universe did not provide a ready-to-use function. Then a more general approach will be handy.


```{r}
iris2 <- iris %>% 
  mutate(iris_sum = pmap(., sum))

iris2 %>% 
  select(iris_sum) %>% pull() %>% (head)
```

Hang on, what data type is this column of?

```{r}
class(iris2$iris_sum)
```

that's a list column. Let's unnest it.


```{r}
iris2 %>% 
  unnest() %>% 
  pull(iris_sum) %>% class()
```

OK.


# Example 2

Let's compute the `rowMeans` in a similar way. One compliation is that  `mean` does not take dots as input - as `sum` does - but takes vector input. Hence, we need to change the domain of the fucntion from vector to dots (list), that's what `lift_vd` does.



```{r}
iris2 <- iris %>% 
  mutate(iris_mean = pmap(., lift_vd(mean))) %>% 
  unnest()

iris2 %>% 
  select(iris_mean) %>% 
  head()
```


# Bonus

Instead of `pull` there's also `pluck` in `purrr`:


```{r}
iris2 %>% pluck("iris_mean")
```


