---
title: Why standard regression is not (so) adequate for regressing proportions
author: Sebastian Sauer
date: '2019-01-03'
slug: why-standard-regression-is-not-so-adequate-for-regressing-proportions
categories:
  - rstats
tags: [tutorial]
editor_options: 
  chunk_output_type: console
---


# Intro

Professor Sweet is conducting some research to investigate the risk factor and drivers of student exam success. In a recent analysis he considers the variable "exam sucessfully passed" (vs. not passed) as the criterion (output) and the amount of time spent for preparation (aka study time) as predictor.


```{r knitrsetup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp =  0.618ß,  #0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Setup

```{r message = FALSE}
data(stats_test, package = "pradadata")
library(tidyverse)
library(ggpubr)
library(glue)
library(ggExtra)
library(broom)
```



Let's change some output formatting of ggplot (see [this post](https://data-se.netlify.com/2018/12/12/changing-the-default-color-scheme-in-ggplot2/) for details)

```{r}
options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill = "viridis")
scale_colour_discrete <- scale_colour_viridis_d
scale_fill_discrete <- scale_fill_viridis_d

theme_set(theme_pubr())

```


# Glimpse

First, Professor Sweet checks the univariate distribution of each variable of interest:

```{r}
stats_test <- stats_test %>% 
  mutate_if(is.integer, as.numeric)

glimpse(stats_test)
```


# Add some noise

Let's add some noise to study time in order to adjust for measurement error:


```{r}
stats_test <- stats_test %>% 
  mutate(study_time_noise = study_time + rnorm(n = nrow(.), 0,.5))
```



```{r}
p1 <- stats_test %>% 
  select(study_time_noise, score) %>% 
  map2(., names(.), ~ggplot(data = data_frame(.x), aes(x = .x)) + 
         labs(x = .y,
              title = glue("Variable: {.y}")) + 
         geom_histogram())

print(p1)
```



# Bivariate association

```{r}
p2 <- ggplot(stats_test) +
  aes(x = study_time_noise, y = score) +
  geom_jitter(alpha = .5) +
  geom_smooth() +
  labs(title = "Bivariate association of study time \nand score")

p3 <- ggMarginal(p2)
p3
```



# Regression

Now, Professor Sweet computes a simple regression. Curiously, he sits and watches what the software spits out. First, a visual sketch:

```{r}
stats_test %>% 
  ggplot() +
  aes(y = score, x = study_time_noise) +
  geom_jitter(alpha = .5) +
  geom_smooth(method = "lm")
```



Then he computes the actual regression:


```{r}
lm1 <- lm(score ~ study_time_noise, data = stats_test)
  
tidy(lm1)  
```


Compute the predicted values:

```{r}
stats_test <- stats_test %>% 
  mutate(pred_lm1 = predict(lm1, newdata = stats_test))
```


Visual sketch:

```{r}
p4 <- stats_test %>% 
  ggplot() +
  aes(x = study_time_noise) +
  geom_jitter(aes(y = score), alpha = .5) +
  geom_line(aes(y = pred_lm1), color = "red", size = 5) +
  labs(title = "Regression of exam score \nby study time")

p4
```



But what would happen if study time was, say, 10?


```{r}
data_new <- data_frame(
  study_time_noise = 0:10,
  score = predict(lm1, newdata = data.frame(study_time_noise = 0:10))
)
```



```{r}
p5 <- ggplot(stats_test) +
  aes(x = study_time_noise, y = score) +
  geom_jitter(alpha = .5) +
  geom_line(data = data_new, color = "red", size = 3) +
  scale_x_continuous(limits = c(0,10)) +
  scale_y_continuous(breaks = seq(0,1,.25)) +
  labs(title = "Regression with high values \nof study time")


p5
```



# Pause

Now the exhausted professor sits and pauses in an attempt to conceive whats has been achieved.


```{r}
ggarrange(p3, p4, p5, labels = c("A", "B", "C"))
```



This approach may yield implausible score values, ie above 1. Clearly, a different approach is needed.



# Sigmoid curve

A simple alternative would be to replace to straight line by a sigmoid curve, such as in the following.

This function computes it:

```{r}
glm1 <- glm(score ~ study_time_noise, data = stats_test, family = "binomial")
```


Add the according predicted values:

```{r}
stats_test <-  stats_test %>% 
  mutate(pred_glm1 = predict(glm1, newdata = stats_test, type = "response"))
```



Plot the predicted values:


```{r}
df_glm <- data_frame(
  study_time_noise = 0:10,
  pred_glm1 = predict(glm1, 
                      newdata = data.frame(study_time_noise = 0:10),
                      type = "response")
)
```



```{r}
p6 <- ggplot(stats_test) +
  aes(x = study_time_noise, y = score) +
  geom_jitter(alpha = .5) +
  labs(title = "Sigmoid regression line") + 
  geom_line(data = df_glm, aes(y = pred_glm1), 
               color = "red", size = 3)

p6
```




# Pause again

```{r}
ggarrange(p3, p4, p5, p6, labels = c("A", "B", "C", "D"))
```

