---
title: Ornaments with ggformula
author: Sebastian Sauer
date: '2019-02-12'
slug: ornaments-with-gformula
categories:
  - rstats
tags:
  - visualization
  - plotting
  - ggplot2
  - tidyverse
editor_options: 
  chunk_output_type: console
---



Since some time, there's a wrapper for `ggplot2` available, bundled in the package `ggformula`. One nice thing is that in that it plays nicely with the popular R package `mosaic`. `mosaic` provides some useful functions for modeling along with a tamed and consistent syntax. In this post, we will discuss some "ornaments", that is, some details of beautification of a plot. I confess that every one will deem it central, but in some cases in comes in handy to know how to "refine" a plot using `ggformula`. 

Note that this "refinement" is primarily controlled via the function `gf_refine()` (most stuff), `gf_lab()` (for labs), and `gf_lims()` (for axis limits). Themes can be adjusted using `gf_theme()`.



# Setup


```{r message = FALSE}
library(mosaic)
library(ggthemes)

data(mtcars)
data(diamonds)
```


# Axis labels


```{r}
gf_point(mpg ~ hp, data = mtcars)  %>% 
  gf_labs(x = "Horsepower",
          y = "Miles per gallone",
          title = "A cool plot")
```






# Axis limits


```{r}
gf_point(mpg ~ hp, data = mtcars) %>% 
  gf_lims(x = c(100,150))
```

Beware! This function will not only "zoom in" but will also kick out the non-displayed data thereby possibly altering geoms such as boxplot or regression line. To just zoom in, use this function instead:


```{r}
gf_point(mpg ~ hp, data = mtcars) %>% 
  gf_refine(coord_cartesian(xlim = c(100, 150)))
```



# Axis breaks


```{r}
gf_point(mpg ~ hp, data = mtcars) %>% 
  gf_refine(scale_x_continuous(breaks = seq(from = 50, to = 400, by = 50)))
```

Basically, `gf_refine()` takes some scaling functions from ggplot as input, such as `scale_XXX_continuous`. So check out these functions to get fine control over axis and, more generally, mapping from variables to visuals.



# Axis labels for discrete axes

```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
  gf_refine(scale_x_discrete(labels  = c("manual", "automatic")))
```

As now the x-axis is discrete (as a result of a factor variables on the x-axis), we need to use `scale_XXX_discrete` this time, where `XXX` is the x-axis in this example.



# Change theme


```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
  gf_theme(theme_classic())
```

```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
  gf_theme(theme_minimal())
```

```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
  gf_theme(theme_bw())
```


```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
  gf_theme(theme_void())
```


# Theme packages


There are some packages out there providing additional themes out of the box. One example is the package `ggthemes`.


```{r}
gf_boxplot(hp ~ factor(am), data = mtcars) %>% 
gf_theme(theme_tufte())
```


# Change colors -- predefined colors


The package `viridis` provides a nice color scheme.

Standard colors:

```{r}
gf_hex(price ~ carat, data = diamonds)
```


```{r}
gf_hex(price ~ carat, data = diamonds) %>% 
  gf_refine(scale_fill_viridis_c()) %>% 
  gf_theme(theme_minimal())
```



# Change colors -- own color scale

I recommend against using own color schemes, at least not without taking great care.


```{r}
gf_boxplot(hp ~ factor(am), 
           data = mtcars,
           fill = ~ factor(am)) %>% 
  gf_theme(theme_classic()) %>% 
  gf_refine(scale_fill_manual(values = c("red", "blue")))
```


Use hex (RGB) color codes instead of the color names:

```{r}
gf_boxplot(hp ~ factor(am), 
           data = mtcars,
           fill = ~ factor(am)) %>% 
  gf_theme(theme_classic()) %>% 
  gf_refine(scale_fill_manual(values = c("#00FF11", "#123456"))) %>% 
  gf_refine(scale_x_discrete(labels = c("Some cars", "some other cars")),
            scale_y_continuous(breaks = c(70, 100, 300)))
```

