---
title: Plotting functions in 3d
author: Sebastian Sauer
date: '2019-11-19'
slug: plotting-functions-in-3d
categories:
  - rstats
tags:
  - plotting
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(mosaic)
library(plotly)
```



# Gimme a function

Say, you have some function such as

$$
f(x) = x^2+z^2
$$


In more R-ish:

```{r}
f <- makeFun(x^2 + z^2 ~ x & z)
```



And you would like to plot it.

Observe that this function has *two* input (independent) variables, $x$ and $z$, plus one output (dependent) variables, $y$.


The thing is, you'll need to define the values for a number of output values for $y$, as defined by the function.


# Define a grid 

First, we define a grid for the input variables.

```{r}
x <- y <- seq(-5, 5, length = 20)
```


# Compute the function value for each combination of the input variables

Then we compute the $y$ value for each combination of $x$ and $y$, this can be achieved using `outer()`. That function applies some function on each combination of $x$ and $y$, and that's just what we would like to achieve.


```{r}
z <- outer(x,y, FUN = f)
```


# Plot in Base R

Now, let's plot:

```{r}
persp(x, y, z)
```



# Plot using Plotly


```{r}
plot_ly(x = x, y = y, z = z, type = "surface")
```

# Bonus: Contour plot in plotly


```{r}
plot_ly(x = x, y = y, z = z, type = "contour")
```




