---
title: Some ways for plotting 3D linear models
author: Sebastian Sauer
draft: FALSE
date: '2019-10-21'
slug: some-ways-for-plotting-3d-linear-models
categories:
  - rstats
tags:
  - regression
---



```{r knitr-setup-plot3d, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(mosaic)
library(plotly)
library(scatterplot3d)
library(rsm)
```


# Motivation

Linear models are a standard way of predicting or explaining some data. Visualizing data is not only of didactical value but provides heuristical value too, as demonstrated by [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet).

Visualizing linear models in 2D is straightforward, but visualizing linear models with more than one predictor is much less so. The aim of this post is to demonstrate some ways do visualize linear models with more than one predictor, using popular R packages. We will focus on 3D examples, that is, two predictors.



# Setup

Some data

```{r}
data(mtcars)
```


Some packages

```{r}
library(tidyverse)
library(mosaic)
```


# Baseline model

For the sake of simplicity, let's start with a 2D model.

```{r}
lm0 <- lm(mpg ~ hp, data = mtcars)
```


## `mosaic``


```{r}
plotModel(lm0)
```


## `ggformula``

```{r}
gf_point(mpg ~ hp, data = mtcars) %>% 
  gf_lm()
```


## `ggplot2`



```{r}
f0 <- function(x) {coef(lm0)[1] + coef(lm0)[2] * x}

ggplot(mtcars) +
  aes(x = hp, y = mpg) +
  geom_point() +
  stat_function(fun = f0)
```


## `makeFun`

Extracting the function itself from a linear model can be achieved like this:

```{r}
f0 <- makeFun(lm0)
f0
```


Then again:


```{r}
ggplot(mtcars) +
  aes(x = hp, y = mpg) +
  geom_point() +
  stat_function(fun = f0)
```



# Our (initial) linear model for 3D


```{r}
lm1 <- lm(mpg ~ hp * disp, data = mtcars)
```


Note that this notation is synonymous to this one:


```{r}
lm1 <- lm(mpg ~ hp + disp + hp:disp, data = mtcars)
```



## `gf_function_2d()`

Let's briefly check the range of the predictors:

```{r}
preds <- c("hp", "disp")
```


```{r}
mtcars %>% 
  dplyr::select(one_of(preds)) %>% 
  summarise_all(funs(min, max))
```

We'll use this information to gauge the limits of the following graph:


```{r}
f1 <- makeFun(lm1)

p1 <- gf_function_2d(f1, xlim = c(50, 300), ylim = c(50, 500)) %>% 
  gf_labs(x = "hp", y = "disp") %>% 
  gf_refine(scale_fill_viridis_c())
p1
```




## Computing a grid and plotting it


First we devise seom combinations of hp and disp values. Then we predict the `mpg` values based on those values using `lm1`.
```{r}
hp_disp_grid <- expand_grid(hp = seq(50, 300, by = 10), disp = seq(50, 500, by= 10))


grid2 <- hp_disp_grid %>% 
  mutate(mpg_pred_lm1 = predict(lm1, newdata = data.frame(hp, disp)))
```


Check

```{r}
gf_density(~mpg_pred_lm1, data = grid2)
```

Now add the predicted value to the `mtcars` dataframe too:

```{r}
mtcars <- mtcars %>% 
  mutate(pred_lm1 = predict(lm1))
```



```{r}
ggplot(aes(x = hp, y = disp), data = grid2) +
  geom_raster(aes(fill = mpg_pred_lm1)) +
  scale_fill_viridis_c() +
  geom_point(data = mtcars, aes(color = pred_lm1)) +
  scale_color_viridis_c()
```






## `plotFun` as tile plot (heatmap)


mosaic provides a convenient function to plot linear models with predictors, `plotFun()`. 

```{r}
plotFun(f1(hp, disp) ~ hp & disp)
```




## `plotFun` as 3d surface plot 


```{r}
plotFun(f1(hp, disp) ~ hp & disp, surface = TRUE)
```





## Plotly surface


For plotly surface, we need the data in *wide* format, ie., we need to spread them out, from long (tidy, standard) format to the wide format. In addition, there should be no more name (key) column, and it ought to be a matrix:


```{r}
grid_wide <- grid2 %>% 
  pivot_wider(names_from = disp, values_from = mpg_pred_lm1) %>% 
  select(-1) %>% 
  as.matrix


rownames(grid_wide) <- seq(50, 300, by = 10)
```




```{r}
xaxis <- list(
  title = "hp"
  #range = c(50, 300)
)

yaxis <- list(
  title = "disp"
  #range = c(50, 500)
)

zaxis <- list(
  title = "mpg"
  #range = c(10, 30)
)

p <- plot_ly(z = ~grid_wide, 
             x = seq(50, 300, by = 10),
             y = seq(50, 500, by = 10)) %>% 
  add_surface(
  contours = list(
    z = list(
      show=TRUE,
      usecolormap=TRUE,
      highlightcolor="#ff0000"
      #project=list(z=TRUE)
      )
    )
  ) %>%
    layout(
    title = "lm1 as 3d plane",
    scene = list(
      xaxis = xaxis,
      yaxis = yaxis,
      zaxis = zaxis
    ))

p
```


## `persp()`

```{r}
x <- seq(min(mtcars$hp), max(mtcars$hp), length.out = 25)
y <- seq(min(mtcars$disp), max(mtcars$disp), length.out = 25)
z <- outer(x, y, function(x,y) {predict(lm1, data.frame(hp = x, disp = y))})
persp(x,y,z , theta = c(45, 135, 225, 315), col = "lightblue")
```



## `contour()`


From package `rsm` we can make use of `contour()`:



```{r}
contour(lm1, hp ~ disp, image = TRUE)
contour(lm1, hp ~ disp, image = FALSE)
```




## Scatterplot 3d


First plot the scatterplot: 
```{r}
d <- mtcars %>% 
  select(hp, disp, mpg)
lm1a <- lm(d$mpg ~ d$hp + d$disp)



p3 <- scatterplot3d(d)


```



Now add the regression plane:


```{r error = TRUE}
p3$plane3d(lm1a)
```


For some reason, this code does not run in (my) rmarkdown setup, but it worked (on my machine) in a ordinary R script.






