---
title: A stochastic problem by Warren Buffet solved with simulation
author: Sebastian Sauer
date: '2019-04-04'
slug: a-stochastic-problem-by-warren-buffet-solved-with-simulation
categories:
  - rstats
tags:
  - simulation
  - rstats
editor_options: 
  chunk_output_type: console
---


```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```


A stochastic problem, with application to financial theory. Some say it goes back to Warren Buffett. I relied to my colleague Norman Markgraf, who pointed it out to me.

Assume there are two coins. One is fair, one is loaded. The loaded coin has a bias of 60-40. Now, the question is: How many coin flips do you need to be "sure enough" (say, 95%) that you found the loaded coin?


Let's simulate la chose.

```{r message = FALSE}
library(mosaic)
library(tidyverse)
library(tictoc)
```


Here's the proportion of successes from a coin flip:


```{r}
{do(10) * rflip(n = 10)}[["prop"]]
```



Let's define a function that does the flipping say a 1000 times with a prespecified (`c_coins`) number of coins with a given bias.

```{r}
flip_loaded_coin <- function(n_coins, 
                             repeats = 1000, 
                             prob = .6) {
  
  distrib <- {mosaic::do(repeats) * 
      mosaic::rflip(n = n_coins, prob = prob)}[["prop"]]

}
```


Now let's flip 1 coin, then 2 coins, ..., then 200 coins, reporting the proportion of success each time.

```{r}
tic()
distrib <- 1:200 %>% 
  map_dfc(~ flip_loaded_coin(.x))
toc()
```

Extract the confidence interval for each sample size (1 to 200 in our simulation).

```{r results = "hide"}
distrib_ci <- distrib %>% 
  map_dfr(~ confint(.x)) %>% 
  add_column(ID = 1:nrow(.)) %>% 
  mutate(bias = ifelse(`2.5%` > .5, TRUE, FALSE))
```



Plot it:


```{r out.width="100%"}
distrib_ci %>% 
  ggplot(aes(x = ID)) +
  geom_hline(yintercept = .5, 
             linetype = "dashed", 
             color = "grey60") +
  geom_errorbar(aes(ymin = `2.5%`,
                    ymax = `97.5%`,
                    color = bias)) +
  theme(legend.position = c(.9, .1))
```


It appears that approx 100 coin flips are needed to quity confidently (for whatever that means) distinguish a baised coin (with a bias of .6) from a fair coin.


