---
title: Geoplotting - update to my MODAR-book
author: Sebastian Sauer
date: '2019-07-29'
slug: geoplotting-update-to-my-modar-book
categories:
  - rstats
tags:
  - geoviz
  - spatial
  - plotting
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```


In my [book](https://link.springer.com/book/10.1007%2F978-3-658-21587-3) on modern data analyisis using R, I show some basics of geoplotting. It seems that some software update for the package [simple features](https://github.com/r-spatial/sf) broke my code. So, here 's some update.


# Load packages and data

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(viridis)
library(sf)


data(socec, package = "pradadata")
data(wahlkreise_shp, package = "pradadata")

```


# Check data

```{r}
glimpse(socec)
```

```{r}
glimpse(wahlkreise_shp)
```


Let's pick only a few columns to make life easier:

```{r}
socec_short <- socec %>%  # Spalten wählen und umbenennen
  select(WKR_NAME = V03,
         WKR_NR = V02,
         wegzug = V11, # Wegzugssalde
         migration = V19,  # Migrationsanteil
         auslaender = V08,  # Auslaenderanteil
         alq = V47,  # Arbeitslosenquote
         migration = V19)  # Anteil mit Migrationshintergrund
```



# Join datasets

There are two data sets: One for the socioeconomic indicators, one for the maps. Let's join them.


```{r}
socec_shape <- socec_short %>%  # Plotten (Arbeitslosigkeit)
  left_join(wahlkreise_shp)
```

This step is not strictly needed (I did not do it the book), but it may save some typing.


Now plot it:

```{r}
socec_shape %>% 
  ggplot +
   aes(fill = alq) +
   theme_void() +
   geom_sf(aes(geometry = geometry)) 

socec_shape %>% 
  ggplot +
   aes(fill = migration) +
   scale_fill_viridis() +
   theme_void() +
   geom_sf(aes(geometry = geometry)) 
```



# What went wrong in my book?

- `geom_sf()` needs to be updated to `geom_sf(aes(geometry = geometry))`. Apparently, the data column with the geo data is not found automatically anymore. I'm unsure about the causes.
- Typo on page 219: after `theme_void()` there was a `+` missing


# sessionInfo

```{r}
sessionInfo()
```


