---
title: How to convert raw scores to different types of standardized scores
author: Sebastian Sauer
date: '2019-04-11'
slug: how-to-convert-raw-scores-to-different-types-of-standardized-scores
categories:
  - rstats
tags:
  - psychology
editor_options: 
  chunk_output_type: console
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```


A common undertaking in applied research settings such as in some areas of psychology is to convert a raw score into some type of standardized score such as z-scores.

This post shows a way how to accomplish that.


# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(umittr)
```


# Load some psychometric data


```{r}
data("extra", package = "pradadata")
```


The data can be downloaded [here](https://github.com/sebastiansauer/pradadata).

The dataset shows some data on extraversion (the personality trait) items along with some correlates of extraversion.

Let's have a look.

```{r}
glimpse(extra)
```


Ah, `extra_mean` is our man. That's the raw score. 

# Convert to z-score with some given mean and sd

Say, the manual or some authority holds the view that the mean equals 3 and the sd equals 1. OK, let's get a z-value using these parameters.

Here's the formula for z-values:


$$ z = \frac{x - \mu}{sd}$$

Where x is the vector of scores and $\mu$ is the mean.

```{r}
extra <-
  extra %>% 
  mutate(extra_z = (extra_mean - 3)/1)
```


# Convert to Stanine

The z-scores can be convertes using this scheme to Stanine:

- Stanine 1: -1.75
- stanine 2:  -1.75 to -1.25 
- stanine 3:  -1.25 to -0.75
- stanine 4: -0.75 to -0.25
- stanine 5: -0.25 to 0.25
- stanine 6:  0.25 to 0.7
- stanine 7: 0.75 to 1.25
- stanine 8: 1.25 to 1.5
- stanine 9: above 1.75


Now let's compare the z-score to the ranges of stanine scores. 


```{r}
extra <-
  extra %>% 
drop_na(extra_mean) %>%
  mutate(extra_stanine = case_when(
    extra_z < -1.75 ~ 1,
    extra_z < -1.25 ~ 2,
    extra_z < -0.75 ~ 3,
    extra_z < -0.25 ~ 4,
    extra_z < 0.25 ~ 5,
    extra_z < 0.75 ~ 6,
    extra_z < 1.25 ~ 7,
    extra_z < 1.5 ~ 8,
    TRUE ~ 9
  ))
```


# Convert to percentiles


Note that I here use the quantile of the empirical cumulative distribution of the data given.

That's quite straightforward:

```{r}
extra <- 
  extra %>% 
    mutate(extra_p = percent_rank(extra_mean),
           extra_p_theoretical = pnorm(extra_z))
```

Use `extra_z` to get the theoretically derived percentiles.



# Check the results

```{r}
extra %>% 
  select(extra_mean, extra_z, extra_stanine, extra_p, 
         extra_p_theoretical) %>% 
  head()
```

