---
title: Reducing residual variance in modeling
author: Sebastian Sauer
date: '2019-03-26'
slug: reducing-residual-variance-in-modeling
categories:
  - rstats
tags:
  - tutorial
---


Modeling is a central part not only of statistical inquiry, but also of everyday human sense-making. We use models as metaphors for the world, in a broader sense. Of course, a model that explains the world better (than some other model) is to be preferred, all other things being equal. In this post, we demonstrate that a more "clever" statistical model reduces the residual variance. It should be noted that this "noise reduction" comes at a cost, however: The model gets more complex; there a more parameters in the model.


First, let's load some data and usual packages:

```{r message = FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
library(gridExtra)

data(tips, package = "reshape2")
```


We filter only the extreme groups in order to carve out the effect more lucidly. In addition, let's compute the residuals and the square residuals:


```{r}
tips2 <- filter(tips, size %in% c(1, 6)) %>% 
  mutate(ID = 1:nrow(.),
         tip_resid = tip - mean(tip),
         tips_resid_quad =tip_resid^2) %>% 
  group_by(size) %>% 
  mutate(tip_mean_group = mean(tip))

glimpse(tips2)
```



`tip_mean_group` refers to the mean value of the response variable `tip` for each conditioned group.



Next, we compute a helper data frame that stores the group means only:

```{r}
tips_sum <- tips2 %>% 
  group_by(size) %>% 
  summarise(tip = mean(tip))
```





Now, let's plot:

That's the plot for the *simpler* model without consideration of subgroups. We model the target variable `tip` as the grand mean (of the whole data set). 

```{r}

p2 <- ggplot(tips2) +
  geom_hline(aes(yintercept = mean(tip))) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = tip,
                   yend = mean(tip)),
               color = "grey80") +
  geom_point(aes(x = ID, y = tip)) +
  labs(title = "tip ~ 1")
p2
```


In contrast, consider the residuals when conditioned on the subgroup means:

```{r}
p1 <- ggplot(tips2) +
  geom_hline(data = tips_sum,
            aes(yintercept = tip,
                color = factor(size))) +
  geom_segment(aes(x = ID,
                   xend = ID,
                   y = tip,
                   yend = tip_mean_group),
               color = "grey80") +
  geom_point(aes(x = ID, y = tip,
                 color = factor(size))) +
  theme(legend.position = "none") +
  labs(title = "tip ~ size")
p1
```



As can be seen, the residuals become *smaller* in this model: The more complex model, where the target variable is modeled as the subgroup average, explains the data better. This models shows a better fit to the data.

```{r out.width="100%"}


grid.arrange(p2, p1, nrow = 1)
```




