---
title: Online reaction time experiments using lab.js
author: Sebastian Sauer
date: '2019-01-29'
slug: online-reaction-time-experiments-using-lab-js
categories:
  - research
tags:
  - research
  - psychology
  - psychometrics
---


Collecting data over the internet used to be fancy, some twenty years or so ago. Nowadays it can be considered standard, if not old school (collecting data using mobile apps is where the cool kids go at the moment). However, there's one noteable exception: Collecting reaction time data over the internet remained a challenge. The reason is simply a technological artefact in that an html response time may vary, vary too much as  to invalidate the signal from some behavorial reaction time research study.


```{r echo=FALSE}
blogdown::shortcode('tweet', '905396423570726912')
```



There are some software packages outthere, such as [eprime](https://pstnet.com/products/e-prime/), probably the most well known software package for reaction time studies, good, but co$stly. Same argument applies for [inquisit](https://www.millisecond.com/), another nice software package, that costs quite a bit.

Of course, that good stuff has its price has never been devil's work, but there are other options available by now. Most noteably, there's some *free* and *open access* tool by the name of [lab.js](https://labjs.felixhenninger.com/). By virtue of its free nature, I've been exciting to getting to know to it. While it's still not as matured as the commercial counterparts, it readily offers a great variety of reaction time based study.



```{r echo=FALSE}
blogdown::shortcode('tweet', '1074711975169540096')
```

One great advantage over the other software packages is that it does *not* need the participants to download some progrm in order to run the experiment - as is needed by inquisit for instance. Alle the experiment is conducted by your browser. That's possible  because lab.js is built on Java Script, which every modern browsers understands. Just awesome!

Download my demo study [here](https://data-se.netlify.com/download/Necker_stuy_labjs.json).

lab.js produces some html files along with some JS files; upload these files to a server (you'll need writing access), and that's it. A number of hosting options for your study are possible; of course, as said, you can always put the files on your own web server.

To get kickstarted, I suggest you read the gread [docs](https://labjs.readthedocs.io/en/latest/learn/builder/). There's also a paper on [psycarxiv](https://psyarxiv.com/fqr49/).




