---
title: Barplots with mosaic
author: Sebastian Sauer
date: '2019-01-10'
slug: barplots-with-mosaic
categories: []
tags:
  - rstats
  - plotting
editor_options: 
  chunk_output_type: console
---


Plotting barplots is a frequent endeavor for the analysis of qualitative data. Numerous methods for plotting barplots exist; the popular R package `mosaic` also provides methods.

More recently, `mosaic` switched to a ggplot wrapper for plotting diagrams, that is `gf_XXX()`, packaged in `ggformula`. That implies that input data is expected to be tidy, because ggplot, a central member of the tidyverse, excepts its input data to be tidy.

Let's check an example.

# Setup

```{r message  = FALSE}
library(mosaic)
data(tips, package = "reshape2")
```


# Old way of barplots in mosaic


The "old" way of plotting barplots in mosaic was lattice based:

```{r}
bargraph(~day, data = tips)
```




# Tidyverse, simple

Here, the present day tidyverse style:


```{r}
gf_bar( ~ day, data = tips)
```


That's simple.


# More complex case, dplyr-style


Assume we are doing some analysis upfront, and we do it the dplyr way:

```{r}
tips_count <- tips %>% 
  filter(tip > 2, day != "Thu") %>% 
  count(day)

tips_count
```

That's a tidy output. Now again, let's plot the frequencies:

```{r}
gf_col(n ~ day, data = tips_count)
```

Simple, works.


# More complex case, using mosaic


```{r}
tips_count2 <- tips_count <- tips %>% 
  filter(tip > 2, day != "Thu") %>% 
  mosaic::tally(~day, data = .)

tips_count2
str(tips_count2)
```

mosaic's `tally()` does not produce a tidy data frame, so cannot (directly) be handed over to `gf_XXX()`.


# Workarounds

Of course, there are workarounds, if one wants to go on using mosaic's `tally()`. For instance, gathering works:


```{r message = FALSE, warning = FALSE}
library(tidyverse)
tips_count2 %>% 
  as_tibble() %>% 
  tidyr::gather(key = day, value = n)
```


Tidy again. Note that `gather()` is not attached by `mosaic()`, hence the call `library(tidyverse)`.


Again, there are functions which are happy to digest a `table` (aka, not tidy) as given back by `mosaic::tally()`:

```{r}
barplot(tips_count2)
```


# `df_stats()` does not work, apparently

[From the package description](https://cran.r-project.org/web/packages/mosaicCore/mosaicCore.pdf) of `mosaicCore` for `df_stats()`:

>   Creates a data frame of statistics calculated on one variable, possibly for each group formed by
combinations of additional variables. The resulting data frame has one column for each of the
statistics requested as well as columns for any grouping variables.


So, `df_stats()` should be useful to get tidy data. However, for `tally()` it appears not to do so:

```{r}
df_stats(~day, data = tips, tally)

df_stats(~day, data = tips, tally) %>% str()
```


The function outputs a data.frame, but not in tidy format.

Similarly, with LHS formula:

```{r}
df_stats(day ~ sex, data = tips, tally)
```


Some interesting (minor?) differences:

```{r}
df_stats(day ~ sex, data = tips, counts)
df_stats(day ~ sex, data = tips, count)
df_stats(sex ~ day, data = tips, counts)
df_stats(~ day | sex, data = tips, counts, props)
```

Nice, but still not tidy.


# `df_stats()`, long format, one variable - does not work

I just discovered that there's a long format parameter in `df_stats(). Let's checkout that.


```{r}
df_stats( ~ day, data = tips, tally, format = "long")
```

Well, that's kind of tidy, but not really, it seems hard to make use of that. Compare to standard  `tally()` once more:

```{r}
tally(~day, data = tips)
```


# `df_stats()`, long format, two variables - works

Maybe this way:

```{r}
df_stats(~ day | sex, data = tips, tally, format = "long")
count(tips, day, sex)
```

Hey, now it works! We've received a tidy data frame.



# Conclusions

`mosaic` gives tidy data frames using the function `df_stats()`, that works great for numeric summaries, such as `favstats()`. For qualitative variables, things appear more complicated. The base line is that `df_bar()` will work, as the raw data is (practically in most times) tidy. If we want to compute the counts by hand, the trouble starts. With two variables, ie., `day ~ sex` things do workout, but apparently not for simply `~ day`. What to do in this case`?


One conclusion would be not to use `mosaic::tally()` but `dplyr::count()` instead, as `count()` yields back a tidy data frame.

Alternative, the output of mosaic's `tally()` can be "tidyfied" after the fact using `tidyr::gather()`.

The mosaic functions such as `df_stats()` are in active development, so we can expect new developments. In addition, I may be missing something.

