---
title: Mean of the upper half of a Gaussian
author: Sebastian Sauer
date: '2020-07-22'
slug: mean-of-the-upper-half-of-a-gaussian
categories:
  - rstats
tags:
  - simulation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(lsr)
```



# Motivation

Recently, I listened to the great Paul Meehl in the audioscripts of some lectures of him. There, he asked the students 

>   what's the mean value of the upper half of a Gaussian distribution?


Let's explore that using simulation techniques.


# Simulation time

Let's draw some instances from a standard Normal distribution, $X$.


```{r}
n <- 1e05
x <- rnorm(n)
```


Mean and SD in our sample are quite close to what can be expected:


```{r}
mean(x)
sd(x)
aad(x)
```

Note that the aad is 0.8 in a standard normal distribution.

# Upper half


Let's put the data into a tibble.

```{r}
df <- tibble(x = x,
             upper_half = ifelse(x >= 0, TRUE, FALSE))
```



```{r}
df %>% 
  group_by(upper_half) %>% 
  summarise(avg_half = mean(x),
            sd_half = sd(x),
            var_half = var(x),
            md_half = median(x),
            iqr_half = IQR(x),
            aad_half = aad(x),
            n_hlaf = n())
```


As can be seen the average in the halves is about 0.8 standard units away from the cut point (zero).



# Some theory: Half normal distribution

The [half-normal distribution](https://en.wikipedia.org/wiki/Half-normal_distribution) has this expected value: $E[Y]=\mu ={\frac {\sigma {\sqrt {2}}}{\sqrt {\pi }}}$.

From the Wikipedia page:

![](https://en.wikipedia.org/wiki/Half-normal_distribution#/media/File:Half_normal_pdf.svg)



In more simple terms, we have this expected value:

```{r}

E = 1 * sqrt(2) / sqrt(pi)
E
```

In other words, about 0.8 standard units, just as we saw above.

Note that `1` refers to the variance of $X$ which is 1.



Regarding the variance, according to the Wikipedia page sited above, $(Y)=\sigma ^{2}\left(1-{\frac {2}{\pi }}\right)$

```{r}
Var = 1 * (1 - (2/pi))
Var

```


Note that this parameter is proportional to the Variance of $X$.

This value does match nicely the simulated value from above. 

However, according to [Math Wolfram](https://reference.wolfram.com/language/ref/HalfNormalDistribution.html),
$V = \frac{-2 + \pi}{2 \theta^2}$.

Wikipedia states that Wolfram Math uses this parametrization:

$$\theta = \sqrt{\pi/2}$$.

That is,

```{r}
theta = sqrt(pi/2)
theta
```


```{r}
Var2 = (-2+pi) / (2 * theta^2)
Var2
```


# Parameter estimation

According to [Wikipedia](https://en.wikipedia.org/wiki/Half-normal_distribution), the $\sigma$ can be estimated as such:


```{r}
sigma_hat = sqrt((1/n) * sum(x^2))
sigma_hat
```



