---
title: simulating values according to some distribution
author: Sebastian Sauer
date: '2020-06-05'
slug: simulating-values-according-to-some-distribution
categories:
  - rstats
tags:
  - simulation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(mosaic)
```


# What's a Monte Carlo simulation?

A [Monte Carlo Simulation](https://en.wikipedia.org/wiki/Monte_Carlo_method) is a numeric approach to solving difficult problems. Instead of having an analytic way of solving the problem, one just says "ok, let's try it out and see what happens".


# Coin flip distribution


Simalatin a single coin flip (Bernoulli) distribution can be achieved like this:


```{r}
rflip()
```


# Normal distribution

```{r}
rnorm(n = 10, mean = 0, sd = 1)
```




# And so on


For example, the log-norm distribution

```{r}
rlnorm(n = 10)
```



# Sampling distribution


Let's draw 1000 samples from a (standard) Normal of size $n=30$.

```{r}
samples1 <- do(1000) * rnorm(30)
samples1
```

It's probably sufficient to store only the means:

```{r}
samples2 <- do(1000) * (rnorm(30) %>% mean())
samples2
```


And plot it:

```{r}
gf_histogram( ~ result, data = samples2)
```



# Practical example

Here's a dataset with the height and sex (and shoe size) of some students.

```{r}
data(wo_men, package = "pradadata")
```

According to this [source](https://en.wikipedia.org/wiki/Average_human_height_by_country), the average height among German men is about 1.75m. Unfortunately, there is no SD given, so we estimate from the SD from some other countries, at 0.07m (7cm).

Now, if our sample were drawn from this "typial German" distribution, how likely would it be to observe our empirical value?


```{r}
wo_men %>% 
  filter(sex == "man") %>% 
  drop_na(height) %>% 
  summarise(height_avg = mean(height),
            n_men = n())
```

It may appear silly to draw samples in this case where the distribution is so simple and well-known. That's true. However, in more complex scenarios -- such as multivariate Normals -- it's a different story. What's more, in some cases, we will not know the shape of the distribution, but we may well draw samples form it.


```{r}
samples3 <- do(1000) * mean(rnorm(n = 18, mean = 175.4, sd = 7))
```


```{r}
gf_histogram(~ mean, data = samples3) %>% 
  gf_vline(xintercept = 183.11)
```


As can be seen, our value of 183cm occurs infrequently, hence we are in a position to reject the null.



