---
title: Causal language appears to be wide-spread in psychological journals, even if
  not warranted
author: Sebastian Sauer
draft: TRUE
date: '2020-07-01'
slug: causal-language-appears-to-be-wide-spread-in-psychological-journals-even-if-not-warranted
categories:
  - rstats
tags:
  - causality
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```


# Science is concerned about causality, to a large degree


Science does not learn much from correlations (association). If you disagree, 
explain what we learn from the fact that, say, storks and babies are associated (as the saying goes).

I think that readers will not doubt that we do *not* learn much from this association. Why?
Because we know that the correlation is not "real", it is spurious. 
By "not real" we intuitively grasp that the correlation -- is this case -- is not causal.
That's why we agree (I think) that we do not learn much from the correlation of storks and babies.


# Prediction has its merits

Of course, there *is* some place und utility for correlations, that is, for prediction. 
Prediction has its value and its merits. 
If you disagree, then you would not want to know if I were able to predict tomorrow's Wall Street winners.
However, science is concerned not with prediction, at least not primarily and as an end goad, but with understanding.
Understanding, or synonymous explaining, is, to a large degree, knowing the causes of some phenomenon.
A course may be loosely defined as something that, if I change it, it will bring about some change in the effect.
This definition is not perfect, but is helpful in many situations.




# Everybody loves causality 

I hold that many researchers at least implicitly agree because they would like to understand the "effect" or "influence" of some x on some y. If some researcher claims that x "energizes" y, she too uses causal language.

The problem is that causal factors are hard to uncover.
Nature loves to hide, as we know.
Of course, it's perfectly fine to start somewhere,
and to state, here I am, here's my causal model, 
my evidence is week, but I need to start somewhere.

Similarly, it is fine (and maybe more transparent),
to not make causal claims if the study does not lend itself to it.
In such case, one is well advised to abstain from causal language,
and to seek shelter in non-causal language such as "association", "relation", "is a predictor", "x and y covary", 
and so forth.


# Bad incentives

However, there are incentives to use causal language even it is not warranted.
Your paper may have a higher propensity of getting published (and of getting more citations),
if there are bold claims in it.
At least, that's a temptation that may be difficult to resist.


