---
title: Plotting a correlated bivariate Gaussian
author: Sebastian Sauer
date: '2020-05-30'
slug: plotting-a-correlated-bivariate-gaussian
categories:
  - rstats
tags:
  - plotting
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(rockchalk)
library(MASS)
```



# Defining the data


```{r}
myR <- lazyCor(X = 0.7, d = 2)
mySD <- c(1, 1)
myCov <- lazyCov(Rho = myR, Sd = mySD)
```


```{r}
myR
mySD
myCov
```


# Drawing from the multivariate normal

Let's draw 1000 cases. Met $\mu$ be zero.

```{r}
twogauss <- mvrnorm(n = 1000, 
                    mu = c(0,0), 
                    Sigma = myCov) %>% as_tibble()
```



# Estimate densities

```{r}
mydens <- kde2d(twogauss$V1, twogauss$V2)
```


# Plot it

```{r}
plotly::plot_ly(z = mydens$z, 
                x=mydens$x, 
                y=mydens$y, 
                type = "surface")
```



# Bonus: Uncorrelated bivariate Gaussian

This is simpler.

First define a grid for the value range of $x$ and $y$.

```{r}
x <- y <- seq(-6, 6, .1)
```


Define the density function (here Gaussian)

Say, $\sigma=2$.

```{r}
s <- 2
f <- function(x,y){exp(-(y^2+x^2)/(2*s^2))}
```


Compute density for each x-y combination. That is, apply function `f` on each x-y combination. `outer` will do that for us.

```{r}
z <- outer(x,y,FUN = f)
```


Plot it:

```{r}
plotly::plot_ly(z = z, x=x, y=y, type = "surface")
```



# Simpler (and maybe more appropriate) alternative

Plotting in 3D may appear cool, but a simple heat map may be more appropriate.

Check this out:


First, let's come up with some noisy data:

```{r}
myR <- lazyCor(X = 0.3, d = 2)
mySD <- c(1, 1)
myCov <- lazyCov(Rho = myR, Sd = mySD)

twogauss <- mvrnorm(n = 1000, 
                    mu = c(0,0), 
                    Sigma = myCov) %>% as_tibble()

head(twogauss)
```



```{r}

myplot1 <- ggplot(twogauss, aes(x = V1, y = V2))

myplot1 + geom_bin2d() + scale_fill_viridis_c()
```

