---
title: Birth Weight Paradoxon -- A simulation for intuition
author: Sebastian Sauer
draft: TRUE
date: '2020-06-11'
slug: birth-weight-paradoxon-a-simulation-for-intuition
categories:
  - rstats
tags:
  - simulation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(ggdag)
library(gt)
library(dagitty)
```


# The Birth Weight Paradoxon

The Birth Weight Paradoxon have kept on puzzling researchers for decades, as [Pearl notes here]().

In [Wikipedia], this paradox is characterized as such:


>    Traditionally, babies weighing less than a certain amount (which varies between countries) have been classified as having low birth weight. In a given population, low birth weight babies have a significantly higher mortality rate than others; thus, populations with a higher rate of low birth weights typically also have higher rates of child mortality than other populations.

>    Based on prior research, the children of smoking mothers are more likely to be of low birth weight than children of non-smoking mothers. Thus, by extension the child mortality rate should be higher among children of smoking mothers. So it is a surprising real-world observation that low birth weight babies of smoking mothers have a lower child mortality than low birth weight babies of non-smokers.


# Birth Weight Paradoxon through the lens of causal inference


Pearl proposes the following DAG to explain the seeming paradox; 
in its essence, it can be described as a collider bias effect.

Consider this DAG, taken from [Pearl's Book of Why]():


```{r}
dag1_str <- '
dag {
smoke [exposure]
bw [outcome]
smoke -> bw 
defect -> bw
bw -> mort
smoke -> mort
defect -> mort
}'
```


```{r}
dag1 <- dag1_str %>% dagitty() %>% tidy_dagitty()
dag1 
```


```{r}
ggdag(dag1) + theme_dag_blank()
```



```{r}
ggdag_adjustment_set(dag1, node_size = 14) + 
  theme_dag(legend.position = "bottom")
```

If no variable is controlled for, then the association of `smoke` to `bw` shows the true causal effect (given this DAG).

Let's check the causal effect from `smoke` with `mort` as the outcome variable.


```{r}
dag1 %>% 
  ggdag_adjustment_set(exposure = "smoke", outcome = "mort") +
  theme_dag()
```

Again, the backdoors are closed. The association of `smoke` with `mort` denotes the true causal effect (given this DAG).


# Controlling for birht weight

Assume a study that solely includes baby with low birth weights. In other words, this study stratifies on `bw`, or controls for `bw`. 

Let's check the consequences for identifying the causal effect from smoke to mortality.




```{r}
dag1 %>% 
  ggdag_dseparated(from = "smoke", to = "mort", controlling_for = "bw")
```


```{r}
dag_paths(dag1, from = "smoke", to = "mort", 
          adjust_for = "bw",
          paths_only = TRUE) 
```

```{r}
ggdag_paths(dag1, from = "smoke", to = "mort",
            shadow = FALSE) +
  theme(legend.position = "bottom",)
```


```{r}
node_dconnected(dag1)
```


