---
title: 'Simulating data for a Gamma regression '
author: Sebastian Sauer
date: '2020-06-17'
draft: FALSE
slug: simulating-data-for-a-gamma-regression
categories:
  - rstats
tags:
  - modelling
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```



# Intro

A Gamma distribution is useful for modeling positive, right skewed data such as waiting times; it is a continuous function.


In this post, we'll illustrate some properties of the Gamma distribution by simulating a toy example.



# Simulate data and define structural model


Let $X$ be a discrete variable following uniform distribution, and  $x_i \in \{1,2,3\}$.

```{r}
set.seed(42)

n <- 1000
X <- sample(x = c(1,2,3), size = n, replace = TRUE)

hist(X)
```


Let $y_i = 0.1 + 1x_i$, by extension let $\hat{y_i} = 0.1 + 0.1x_i + \epsilon_i$, where $\epsilon_i \sim \Gamma(m = 0.1, \, s= 1/m)$, and $\Gamma$ refers to the Gamma distribution with location $m$ and shape $s$.


```{r}
e <- rgamma(n, shape = 0.1, scale = 1/0.1)
```


Compute $Y$ and $\hat(Y)$:


```{r}
Y = 0.1 + 1 * X
```


```{r}
Y_hat = Y + e
```


```{r}
hist(Y_hat)
```


```{r}
plot(X, Y_hat)
```


# Regression


```{r}
glm1 <- glm(Y_hat ~ X, Gamma(link = "log"))
summary(glm1)
```

# Coefficients, delogged

```{r}
exp(coef(glm1))
```


The delogged (exponentiated) coefficients provide a *factor* by which the mean of the outcome variable will change per unit change of X, similar to a normal linear model. The main difference is that the the coefficient denotes a *multiplicative* change rather than an *additive* change.



# Prediction time

We'd expect that, if $x_i=1$, that $y_i=1.72 * 1.34 \approx 2.31$.

Let's check that.


```{r}
predict(glm1, newdata = data.frame(X=1))
```


That's in log-notation. Let's delog:


```{r}
predict(glm1, newdata = data.frame(X=1)) %>% exp()
```


Works!
