---
title: Plot for mean comparison
author: Sebastian Sauer
date: '2020-06-02'
slug: plot-for-mean-comparison
categories:
  - rstats
tags:
  - plotting
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(reshape2)  # for data
library(mosaic)
library(sjmisc)
library(skimr)
```


# Data setup

```{r}
data(tips)
```


# Aggregate data per group


```{r}
tips_aggr <- tips %>% 
  group_by(smoker) %>% 
  summarise(tip_avg = mean(tip),
            tip_md = median(tip),
            tip_sd = sd(tip),
            tip_iqr = IQR(tip))

tips_aggr
```

The same lines, more concisely:

```{r}
tips_descr <- tips %>% 
  group_by(smoker) %>% 
  descr(tip)

tips_descr
```

`descr` handles back a list, which may be not practical for further processing. 

`skim` provides another alternative:

```{r}
tips_skim <- tips %>% 
  group_by(smoker) %>% 
  skim(tip)

tips_skim
```

This function gives back a tidy data frame. Nice.



Alternative, using `mosaic`:


```{r}
tips_fav <- tips %>% 
  favstats(tip ~ smoker, data = .)
tips_fav
```



# Plot 1


```{r}
ggplot(tips_skim) +
  aes(x = smoker, y = numeric.mean) +
  geom_line(group = 1) +
  geom_pointrange(aes(ymin = numeric.mean - numeric.sd,
                    ymax = numeric.mean + numeric.sd),
                  color = "grey40") +
  geom_point(size = 5) +
  ylim(0, 5) +
  labs(caption = "Error bars represent standard deviation",
       y = "average tip")
  
```




# List to data frame
 
`tips_descr` gives us a list, but more often that not, we would like to go on using a tibble. That's what's  `enframe` is for. Subsequently, we can make use of `unnest` to unnest the list-column `value`. 



```{r}
tips_descr %>% 
  enframe() %>% 
  unnest(value)  # that's the name of the list-column to be unnested
```

