---
title: Chi-squared test using simulation based inference
author: Sebastian Sauer
date: '2020-06-04'
slug: chi-squared-test-using-simulation-based-inference
categories:
  - rstats
tags:
  - simulation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```


# Simulation based inference 

Simulation based inference (SBI) is an elegant way of subsuming a wide array of statistical (inference) methods under one umbrella. In addition, its simple thereby helping learners getting to the grips.


[Here's](https://allendowney.blogspot.com/2016/06/there-is-still-only-one-test.html) a summary of the central ideas.

However, this post does not aim at explaining simulation based inference, which is done [elsewhere](http://mosaic-web.org/go/SM2-technique/).


# Testing the association of two categorical variables

One application of statistical tests -- simulation based or classical -- is testing the association of two categorical variables.

Let's try that using SBI.


# Setup

```{r}
library(mosaic)
data(tips, package = "reshape2")
```


# Research question: Is there an assocation of day and smoker


Let's test whether there's an association of `day` and `smoker`.


First, the sample descriptives:

```{r}
tally( sex ~ day, data = tips, format = "proportion")
```


Interesting: On Saturday on Sunday, mostly men pay. On Friday and Thursday, there's an equity of pay between the sexes. At least in our data. That brings us to the point: How about the population?


# Chi-Square statistic

The $\chi^2$ statistic gives us a measure of deviation from expected value.


```{r}
chi_value <- chisq(sex ~ day, data = tips)
```

Alas, this value is not so intuitive. We can extract the p-value from the theoretical distribution like this:

```{r}
pchisq(chi_value, 3, lower.tail=FALSE)
```

Let's compare it to the result of the $\chi^2$ test:


```{r}
x <- xchisq.test(sex ~ day, data = tips)
```

Which is (approximately) the same.



# $\chi^2$ via SBI

This statistic can easily be submitted to SBI, ie., the permutation test:

```{r}
null1 <- do(1000) * chisq(sex ~ shuffle(day), data = tips)
``` 

Let's check the distribution of the $\chi^2$ statistic:

```{r}
gf_histogram(~ X.squared, data = null1)
```


Now let's add a marker for our empirical value:


```{r}
x$statistic
```


# Plot it

```{r}
gf_histogram(~ X.squared, data = null1) %>%  gf_vline(xintercept = ~ x$statistic)
```

As can easily be seen, the empirical value is an infrequent event in the null distribution.

More precisely:

```{r}
p_value <- prop(~ X.squared >= x$statistic, data = null1)
p_value
```

Incidentally, this simulated p-value is approximately the value we got from the theoretical distribuion.



We might want to add the critical value to this graph:

```{r}
crit_value <- qchisq(.05, df = 3, lower.tail=FALSE)
crit_value
```


`crit_value` gives us the critical value for a $\chi^2$-distribution with $df=3$ degrees of freedom.




```{r}
```{r}
gf_histogram(~ X.squared, data = null1) %>% 
  gf_vline(xintercept = ~ x$statistic) %>% 
  gf_vline(xintercept = ~ crit_value, linetype = "dashed")
```
```




