---
title: Adjustment set exercise from Elwert 2013
author: Sebastian Sauer
date: '2020-05-19'
slug: adjustment-set-exercise-from-elwert-2013
categories:
  - causal
  - DAG
tags:
  - rstats
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(ggdag)
library(dagitty)
```


# Define DAG

I've drawn the DAG in [dagitty.net](http://dagitty.net/dags.html), that's why the coordinates look weird.

```{r}
dag3_str <- '
dag {
bb="-2.865,-5.146,2.956,4.896"
U [latet, pos="2.456,-0.958"]
X [exposure, pos="-2.365,-4.309"]
Y [outcome, pos="-0.271,4.059"]
Z1 [pos="-0.491,-1.925"]
Z2 [pos="-0.915,1.269"]
Z3 [pos="1.713,1.984"]
U -> Z1
U -> Z3
X -> Z1
Z2 -> Y
Z2 -> Z1
Z2 -> Z3
Z3 -> Y
}'
```

Then tidify:

```{r}
dag3 <- dagitty(dag3_str)

dag3_tidy <- tidy_dagitty(dag3)

dag3_tidy
```


# Plot


```{r}
ggdag(dag3_tidy) + theme_dag()
```


# Find adjustment sets

Which paths need be closed in order to get the causal effect of $X$ on $Y$?


```{r}
ggdag_adjustment_set(dag3_tidy, node_size = 14) + 
  theme_dag(legend.position = "bottom")
```


No need to control for any variable!


# Controlling for Z1



```{r}
dag3_tidy %>% 
  ggdag_dseparated(from = "X", to = "Y", controlling_for = "Z1")
```


If we control for $Z1$, a bunch of paths fom $X$ to $Y$ get openened, as can be seen in the DAG:

1. X - Z1 - U - Z3 - Y
2. X - Z1 - Z2 - Y
3. X - Z1 - Z2 - Y


As a consequence, we have open backdoors from $X$ to $Y$. The causal effect is now confounded by these additional back doors.

