---
title: Randomization in presence of an interaction effect
author: Sebastian Sauer
date: '2020-07-07'
slug: randomization-in-presence-of-an-interaction-effect
categories:
  - rstats
tags:
  - simulation
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
library(rockchalk)
library(MASS)
library(ggdag)
```



# Problem statement

Assume that $X$ and $Y$ are correlated contingent on some third variable, $Z$. For simplicity, assume that, if $z=0$, $_0=0.7$, and if $z=1$, then $r_1=-0.7$. This is not a causal statement.


# Simulate data

Let the sample size amount to $n=1000$.

```{r}
n <- 1e03
```



Group A, $z=0$:

```{r}
myR <- lazyCor(X = 0.7, d = 2)
mySD <- c(1, 1)
myCov <- lazyCov(Rho = myR, Sd = mySD)

set.seed(12345)
x_z0 <- MASS::mvrnorm(n=n, mu = rep(0, 2), Sigma = myCov)


head(x_z0)
```



```{r}
cor(x_z0)
```





Group B, $z=1$:

```{r}
myR <- lazyCor(X = -0.7, d = 2)
mySD <- c(1, 1)
myCov <- lazyCov(Rho = myR, Sd = mySD)

set.seed(54321)
x_z1 <- MASS::mvrnorm(n=n, mu = rep(0, 2), Sigma = myCov)


head(x_z1)
```



```{r}
cor(x_z1)
```




# Enframe


```{r}
x_z0_df <- data.frame(x_z0)
x_z1_df <- data.frame(x_z1)



mydata_raw <- bind_rows(x_z0_df, x_z1_df)

mydata <- mydata_raw %>% rename(x = X1, y = X2)

head(mydata)
```

Add group information:


```{r}
mydata <- mydata %>% 
  mutate(z = c(rep(0,n), rep(1, n)))

head(mydata)
```



# Marginal correlation


We know the conditional correlation, because we simulated it. Now let's look at the marginal (unconditional) correlation:



```{r}
mydata %>% 
  dplyr::select(x, y) %>% 
  cor()
```

In sum, we have zero correlation.



# Moderation framework

$Z$ can be seen as a moderator in a regression framework:


```{r}
lm1 <- lm(y ~ x * z, data = mydata)

summary(lm1)
```

We see that $Z$ is itself *not* associated with the outcome, $Y$, but it changes the association of $X$ with $Y$.

```{r}
mydata %>% 
  ggplot(aes(x = x, y = y, color = factor(z))) +
  geom_point() +
  geom_smooth(methode = "lm", se = FALSE)
```


Compare that to this model, which does not take the moderator into account:

```{r}
lm2 <- lm(y ~ x, data = mydata)

summary(lm2)

```


```{r}
ggplot(mydata, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

No (marginal) association, as can be seen in the above diagram.



# Causal explanation


One causal model explaining this interaction could be:


```{r}
dag1 <- dagitty::dagitty("dag{
x -> z
y -> z
}")

dag1
```



```{r}
dag1_tidy <- tidy_dagitty(dag1)
ggdag(dag1) +
  theme_dag_blank()
```


This causal setting is called a *Collider Bias*, see e.g. [this source](https://rdrr.io/cran/ggdag/f/vignettes/intro-to-dags.Rmd).


```{r}
collider_triangle() %>% 
  ggdag_dseparated(controlling_for = "m") +
  theme_dag()
```


Controlling for *m* ($Z$ in our notation) introduces a bias, that is a spurious association between x and y.


# Regression according to Collider model


```{r}
lm_collider <- lm(z ~ x*y, data = mydata)

summary(lm_collider)
```


However, there's no association (marginally) between $X$ and $Y$:

```{r}
lm_xy <- lm(y ~ x, data = mydata)

summary(lm_xy)
```






# Effect of randomization of $Z$

Now let's randomize (permutate) $Z$ to see what happens.


```{r}
mydata <- mydata %>% 
  mutate(z_permutated = sample(z, size = length(z)))

head(mydata)
```



```{r}
mydata %>% 
  dplyr::select(x, y, z, z_permutated) %>% 
  cor()
```



# Unconditional correlated remains unchanged


```{r}
mydata %>% 
  dplyr::select(x, y, z) %>% 
  group_by(z) %>% 
  summarise(r_marginal = cor(x,y))
```



# The permutation deleted the correlation 

Within each group, that is, $z=0$ and $z=1$, there's no correlation left.


```{r}
mydata %>% 
  group_by(z_permutated) %>% 
  summarise(r_marginal = cor(x,y))
```



# Pause and ponder


Permutating (randomizing) a group variable that acts as a moderator *as in the case above* will erase the moderator effect.

This is due to the fact that a randomized experiment will have all the paths going into the to-be randomized variable having cut. This in turn means that the randomized variable, $Z$ will not be biased/influenced by any other variable.




# Can a fork be a moderator?


Assume the following DAG:




```{r}
dag_fork <- dagitty::dagitty("dag{
x <- z
y <- z
}")

dag_fork
```



```{r}
dag_fork_tidy <- tidy_dagitty(dag_fork)
ggdag(dag_fork) +
  theme_dag_blank()
```

# Simulate data accordingly


```{r}
z <- sample(x = c(0,1), size = n, replace = TRUE)


x <- z + 0.3 * rnorm(n = n)
y <- z + 0.3 * rnorm(n = n)
```



# Put the data together


```{r}
mydata2 <- tibble(
  z = z,
  x = x,
  y = y,
  z_permuated = sample(z, size = n)
)
```



# Regression time


```{r}
lm_fork <- lm(y ~ x * z, data = mydata2)

summary(lm_fork)
```

No, it appears that this is not a way a fork can act as a moderator.



# Can achain act as a moderator?


Assume the following DAG:


```{r}
dag_chain <- dagitty::dagitty("dag{
z -> x -> y
}")

dag_chain
```



```{r}
dag_chain_tidy <- tidy_dagitty(dag_chain)
ggdag(dag_chain_tidy) +
  theme_dag_blank()
```


# Simulate data


```{r}
z1 <- sample(c(0,1), size = n, replace = TRUE, prob = c(0.3, .7))
z2 <- sample(c(0,1), size = n, replace = TRUE, prob = c(0.7, .3))


mydata3 <- tibble(
  z = c(z1, z2),
  x = z + 0.1 * rnorm(n=n),
  y = x + 0.1 * rnorm(n=n)
)
```




# Regression time


```{r}
lm_chain <- lm(y ~ x*z, data = mydata3)

summary(lm_chain)
```


Given this DAG, we cannot expect to see a moderator effect.


# Debrief

It appears that a moderator is only compatible with a limited number of causal models. Secondly, we see that randomizing (permutating) erases the association, as should be expected.

There can be said much more to the relation of moderation (interaction) and causality, see eg., [this paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2235194/), or [this](https://academic.oup.com/ije/article/45/6/1904/2670330) or [this](https://journals.lww.com/epidem/Fulltext/2009/11000/Four_Types_of_Effect_Modification__A.16.aspx)


