---
title: How to find the package of a R function
author: Sebastian Sauer
date: '2020-05-15'
slug: how-to-find-the-package-of-a-r-function
categories:
  - rstats
tags:
  - stats
  - tutorial
---



```{r knitr-setup, echo = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold",
  size = "tiny"
)
```



# Load packages

```{r load-libs, message = FALSE, waring = FALSE}
library(tidyverse)
```



# Where does my function reside?


Finding the package of a given R function is some hassle. I am not aware of a quick built-in way in R to find the package of a function.

That's why I came up with my own function, check it out:


## Install package

Speaking of packages of function, that's the package where this function stays:

```{r eval = FALSE}
library(devtools)
install_github("sebastiansauer/prada")
```


## Example

```{r}
library(prada)
```


```{r}
find_funs("select")
```


```{r}
find_funs("tidy")
```


Note that his may take a while (up to a few seconds in some cases).

# Code

```{r}
find_funs
```



Enjoy!

